---
title: "README"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
# we pre-load the model, the fitting time is around 30 minutes
fit <- readRDS(file = "prefitted/Anderson2010.RDS")
```

# Robust Bayesian Meta-Analysis (RoBMA)

This package estimates and ensemble of meta-analytical models (assuming either presence or absence of the effect, heterogeneity, and publication bias) and uses Bayesian model averaging to combine them. The ensemble uses Bayes Factors to test for the presence of absence of the individual components (e.g., effect vs no effect) and model-averages parameter estimates based on posterior model probabilities. The user can define a wide range of non-informative or informative priors for the effect size, heterogeneity, and weight 
functions. The package provides convenient functions for summary, visualizations, and fit diagnostics.

See our pre-print, Maier, Bartoš, & Wagenmakers (2020) at https://doi.org/10.31234/osf.io/u4cns, for more details about the implementation, examples, and simulation studies.

## Installation

The development version can be installed from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("FBartos/RoBMA")
```

In case of compilations problems, please, install the package from source using the following [RoBMA.tar.gz](https://drive.google.com/file/d/11rlNum7YpsjzaF0aRqzi3XjedmRLjwX9/view?usp=sharing)

``` r
# install.packages("devtools")
install.packages("RoBMA.tar.gz", repos = NULL, type="source")
```


## Example

This is a reproduction of an example in Maier, Bartoš, & Wagenmakers where we use the RoBMA package to re-analyze a subset of Anderson et al. (2010) meta-analysis of effects of violent video games on aggression (the original dataset can be found [here](https://github.com/Joe-Hilgard/Anderson-meta)).

First, we load the dataset which is included with the package. The dataset contains correlational coefficients, sample sizes, and labels for 27 experimental studies focusing on the effect of violent videogames on aggressive behavior.

```{r}
library(RoBMA)

data("Anderson2010", package = "RoBMA")
head(Anderson2010)
```

Then, we fit the meta-analytic model ensemble that is composed of 12 models (the default settings of RoBMA fitting function). These models are all possible combinations of priors for the following parameters (the prior parameters for mu and tau are defined on Cohen’s d transformed scale if correlation coefficients are supplied):

* mu (the mean parameter)
  - a spike at zero, representing a null effect
  - a standard normal distribution, representing the alternative hypothesis
* tau (the heterogeneity parameter)
  - a spike at zero, representing no heterogeneity (fixed effect)
  - an inverse gamma distribution with shape = 1 and scale = 0.15, representing presence of heterogeneity (random effects), based on van Erp et al. (2017) 
* omega (the weights parameters for modeling publication bias)
  - a spike at 1, representing no publication bias (all studies have the same probability of being published)
  - a two-steps two-sided weight-function with p-values cut-off at 0.05 and weights following a cumulative sum of Dirichlet distribution with alpha = (1,1)
  - a three-steps two-sided weight-function with p-values cut-offs at 0.05 and 0.10 and weights following a cumulative sum of Dirichlet distribution with alpha = (1,1,1)

The prior odds of these prior distributions are by default set to make all three model categories equally likely a priory (0.5 prior probability of an effect, 0.5 prior probability of heterogeneity, and 0.5 prior probability of publication bias).

The default models using the correlation coefficients can then be fitted using the `RoBMA()` function (all input is internally transformed into t-statistics and degrees of freedom):

``` r
fit <- RoBMA(r = Anderson2010$r, n = Anderson2010$n, study_names = Anderson2010$name)
```

The main model summary can be obtained using the `summary()` function.

The first table shows an overview of the model types -the number of models, prior and posterior probability, and inclusion Bayes factor. As we can see, the data show strong evidence for the presence of effect and publication bias and weak evidence for lack of heterogeneity.

The second table shows model-averaged estimates weighted by the individual models' posterior probabilities. The mean estimate mu (converted back to correlation scale) is considerably lower than in the original meta-analysis due to the publication bias correction. The heterogeneity estimate tau (on the Cohen's d scale) has most of its probability mass around zero due to the higher support of models with no heterogeneity, and the parameters omega, representing the weights at each p-value interval are decreasing with increasing p-values, showing the publication bias.

```{r}
summary(fit)
```

We can visualize the estimated parameters using the `plot()` function. In the case of parameter tau, the arrow stands for probability mass at tau = 0. Weights omega are by default plotted as the weight function, which can be changed to the individual weights estimates by setting `weights = TRUE`.

```{r fig_mu, out.width = '50%', fig.align = "center"}
plot(fit, parameter = "mu")
```

```{r fig_tau, out.width = '50%', fig.align = "center"}
plot(fit, parameter = "tau")
```

```{r fig_omega, out.width = '50%', fig.align = "center"}
plot(fit, parameter = "omega")
```

Furthermore, we can inspect the individual models' estimates mu, including the prior and posterior probability.

```{r fig_mu_ind, fig.height = 12.5, fig.width = 7, out.width = '75%', fig.align = "center"}
plot(fit, parameter = "mu", type = "individual")
```

Apart from plotting, the individual model performance can be inspected using the `summary()` function with argument `type = "individual"`. An overview of the individual model MCMC diagnostics can be obtained by setting `type = "models", diagnostics = TRUE` (not shown here for the lack of space).

We can also visualize the MCMC diagnostics using the diagnostics function. The function can display the chains `type = "chain"` / posterior sample densities `type = "densities"`, and averaged autocorrelations `type = "autocorrelation"`. Here, we request the chains trace plot of the mu parameter of the most complex model by setting `show_models = 12` (the model numbers can be obtained from the summary function with `type = "models"` argument.)

```{r fig_mu_chain, out.width = '50%', fig.align = "center"}
RoBMA::diagnostics(fit, parameter = "mu", type = "chains", show_models = 12)
```

The package allows to fit highly customized models with different prior distribution functions, prior model probabilities, and provides more visualization options. See the documentation to find out more about the specific functions: `?RoBMA`, `?priors`, `?plot.RoBMA`. The main package functionalities are to be implemented within the Meta Analysis module of JASP 0.13.

## Sources
Anderson, C. A., Shibuya, A., Ihori, N., Swing, E. L., Bushman, B. J., Sakamoto, A., ... & Saleem, M. (2010). Violent video game effects on aggression, empathy, and prosocial behavior in Eastern and Western countries: A meta-analytic review. Psychological bulletin, 136(2), 151.

Maier, M., Bartoš, F., & Wagenmakers, E. (2020, June 23). Robust Bayesian meta-analysis: Addressing publication bias with model-averaging. https://doi.org/10.31234/osf.io/u4cns

JASP Team. (2020). JASP (Version 0.13)[Computer software].

van Erp, S., Verhagen, J., Grasman, R. P., & Wagenmakers, E. J. (2017). Estimates of between-study heterogeneity for 705 meta-analyses reported in Psychological Bulletin from 1990–2013. Journal of Open Psychology Data, 5(1).


