---
title: "Bayesian Meta-Analysis"
author: "František Bartoš"
date: "28th of April 2026"
output:
  rmarkdown::html_vignette:
    self_contained: yes
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Bayesian Meta-Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---

```{r setup, include = FALSE}
source("vignette-cache.R", local = knitr::knit_global())

cached_fits <- vignette_cache(
  name     = "02-bayesian-meta-analysis",
  objects  = c("fit1_brma", "fit2_brma", "fit3_brma"),
  packages = c("metafor", "metadat")
)

knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  eval       = vignette_cache_eval(cached_fits),
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 3.5,
  dev        = "png"
)
if (.Platform$OS.type == "windows") {
  knitr::opts_chunk$set(dev.args = list(type = "cairo"))
}
```

```{r load-models, include = FALSE}
library("metafor")
library("RoBMA")

data("dat.bcg", package = "metadat")

dat <- metafor::escalc(
  ai = tpos, bi = tneg, ci = cpos, di = cneg,
  measure = "RR", data = dat.bcg
)

vignette_cache_load(cached_fits)
```

```{r precompute-models, include = FALSE, eval = FALSE}
library("metafor")
library("RoBMA")

data("dat.bcg", package = "metadat")

dat <- metafor::escalc(
  ai = tpos, bi = tneg, ci = cpos, di = cneg,
  measure = "RR", data = dat.bcg
)

fit1_brma <- brma(
  yi = yi, vi = vi, measure = "RR",
  data = dat, seed = 1
)
fit1_brma <- add_loo(fit1_brma)
fit1_brma <- add_marglik(fit1_brma)

fit2_brma <- brma(
  yi = yi, vi = vi, measure = "RR",
  mods = ~ ablat,
  data = dat, seed = 1
)
fit2_brma <- add_loo(fit2_brma)
fit2_brma <- add_marglik(fit2_brma)

fit3_brma <- brma(
  yi = yi, vi = vi, measure = "RR",
  mods = ~ ablat + alloc,
  data = dat, seed = 1
)
fit3_brma <- add_loo(fit3_brma)
fit3_brma <- add_marglik(fit3_brma)

vignette_cache_save(cached_fits)
```

A Bayesian random-effects meta-analysis with `brma()` mirrors `metafor`'s `rma()` in interface: the same effect-size, moderator, and meta-regression specifications drop in unchanged.
The model output is a posterior distribution; credible intervals, Bayes factors, and leave-one-out predictive scores follow from the same fit.

This vignette walks through the BCG vaccine examples from `metafor` [@metafor] alongside their `brma()` counterparts [@RoBMA], one model at a time.
All `brma()` calls keep the default priors; see the *Prior Distributions* vignette for the prior definitions and customization options.
The outcome is the log risk ratio (`measure = "RR"`); for the broader Bayesian framework, see @bartos2025bayesian.

## Random-Effects Model

The BCG vaccine dataset from `metadat` contains 13 randomized trials on tuberculosis prevention.
`escalc()` computes log risk ratios `yi` and their sampling variances `vi`.

```{r escalc}
data("dat.bcg", package = "metadat")

dat <- metafor::escalc(
  ai = tpos, bi = tneg, ci = cpos, di = cneg,
  measure = "RR", data = dat.bcg
)
```

`metafor::rma()` fits the random-effects model.

```{r fit1-metafor}
fit1_metafor <- metafor::rma(yi, vi, data = dat, method = "REML")
fit1_metafor
```

The matching Bayesian fit reuses the same effect-size columns.
`measure = "RR"` tells `brma()` to use the default priors for log risk ratios.
The defaults are weakly informative and provide slight regularization toward zero; see the *Prior Distributions* vignette for the prior definitions and rescaling options.

```{r fit1-brma, eval = FALSE}
fit1_brma <- brma(
  yi = yi, vi = vi, measure = "RR",
  data = dat, seed = 1
)
```

`summary()` reports posterior means, credible intervals, and (suppressed here) MCMC convergence diagnostics.
Convergence diagnostics are included by default; chain-level visual diagnostics are available via `diagnostic_plots_trace()`, `diagnostic_plots_density()`, and `diagnostic_plots_autocorrelation()`.

```{r fit1-brma-summary}
summary(fit1_brma, include_MCMC_diagnostics = FALSE)
```

The posterior mean for the log risk ratio is close to the REML estimate from `metafor`; the slight pull toward zero reflects the default `Normal(0, 1)` prior.
`pooled_effect()` returns the same summary on the natural risk-ratio scale via `transform = "EXP"`:

```{r fit1-brma-pooled}
pooled_effect(fit1_brma, transform = "EXP")
```

`plot()` visualizes the posterior, and `prior = TRUE` overlays the prior so the shift from prior to posterior is visible.

```{r fit1-brma-plot, fig.width = 5, fig.height = 3.5}
plot(fit1_brma, parameter = "mu", prior = TRUE, transform = "EXP")
```

## Heterogeneity

`metafor::confint()` reports profile-likelihood confidence intervals for `tau^2`, `tau`, `I^2`, and `H^2` [@higgins2002quantifying].
`summary_heterogeneity()` is the Bayesian counterpart, returning posterior summaries of the same quantities.

```{r heterogeneity-confint}
confint(fit1_metafor)
```

```{r heterogeneity-bayes}
summary_heterogeneity(fit1_brma)
```

## Funnel Plot

`funnel()` plots observed effect sizes against the fitted sampling distribution.
The `RoBMA` output is intentionally close to `metafor::funnel()`.

```{r funnel-plot, fig.width = 7, fig.height = 3.2}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
metafor::funnel(fit1_metafor, main = "metafor")
funnel(fit1_brma, main = "RoBMA")
par(op)
```

## Meta-Regression with a Continuous Moderator

A continuous moderator is added with the standard formula syntax.
`ablat` is the absolute latitude of the trial, frequently used to explain BCG effect-size variation.

```{r fit2-metafor}
fit2_metafor <- metafor::rma(yi, vi, mods = ~ ablat, data = dat)
fit2_metafor
```

```{r fit2-brma, eval = FALSE}
fit2_brma <- brma(
  yi = yi, vi = vi, measure = "RR",
  mods = ~ ablat,
  data = dat, seed = 1
)
```

```{r fit2-brma-summary}
summary(fit2_brma, include_MCMC_diagnostics = FALSE)
```

`regplot()` produces the familiar bubble plot with the fitted regression line.

```{r regplot-continuous, fig.width = 7, fig.height = 3.2}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
metafor::regplot(fit2_metafor, main = "metafor")
regplot(fit2_brma, main = "RoBMA")
par(op)
```

`predict()` returns posterior summaries for user-specified moderator values.
The `yi` and `vi` columns in `newdata` are required by input validation and are not used for `type = "terms"`.

```{r fit2-predict}
new_data <- data.frame(
  ablat = c(15, 35, 55),
  yi    = 0,
  vi    = 1
)
predict(fit2_brma, newdata = new_data, type = "terms")
```

## Comparing Models with Bayes Factors

Marginal likelihoods enable model comparison via Bayes factors [@kass1995bayes], with no direct analogue in `metafor`.
After attaching marginal likelihoods to both fits with `add_marglik()`, `bf()` returns the Bayes factor for one model against another.

```{r add-marglik, eval = FALSE}
fit1_brma <- add_marglik(fit1_brma)
fit2_brma <- add_marglik(fit2_brma)
```

```{r bayes-factor}
bf(fit2_brma, fit1_brma)
```

The Bayes factor quantifies how much the data shift the relative predictive support between the two models.

## Adding a Categorical Moderator

`alloc` is a three-level factor describing the trial allocation method (random, alternate, systematic).
The same formula syntax extends the meta-regression.

```{r fit3-metafor}
fit3_metafor <- metafor::rma(yi, vi, mods = ~ ablat + alloc, data = dat)
```

```{r fit3-brma, eval = FALSE}
fit3_brma <- brma(
  yi = yi, vi = vi, measure = "RR",
  mods = ~ ablat + alloc,
  data = dat, seed = 1
)
```

```{r fit3-brma-summary}
summary(fit3_brma, include_MCMC_diagnostics = FALSE)
```

`metafor::regplot()` plots categorical moderators by selecting one dummy column.
`RoBMA::regplot()` plots the factor as a single moderator while still adjusting for the other moderators in the fit.

```{r regplot-categorical, fig.width = 7, fig.height = 3.2}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
metafor::regplot(fit3_metafor, mod = "allocrandom", main = "metafor: random dummy")
regplot(fit3_brma, mod = "alloc", main = "RoBMA: alloc factor")
par(op)
```

`marginal_means()` computes estimated marginal means for each moderator term, averaging over the other moderators in the fit.
Continuous moderators (here `ablat`) are evaluated at $-1$, $0$, and $+1$ standard deviations of the predictor; categorical moderators (here `alloc`) at each level.

```{r marginal-means}
emm <- marginal_means(fit3_brma)
summary(emm)
```

```{r marginal-means-plot, fig.width = 5, fig.height = 3.5}
plot(emm, parameter = "alloc")
```

## Model Diagnostics

`metafor::rstudent()` returns frequentist studentized residuals computed by deletion;
`RoBMA::rstudent()` returns LOO-PIT residuals, which propagate posterior uncertainty and leverage through leave-one-out prediction.
The two definitions are comparable but not identical, and both require `add_loo()` on the Bayesian side.

```{r add-loo, eval = FALSE}
fit3_brma <- add_loo(fit3_brma)
```

```{r residuals}
head(rstudent(fit3_metafor)$z)
head(rstudent(fit3_brma)$z)
```

`qqnorm()` checks the residuals against a standard normal reference.

```{r qqplot, fig.width = 7, fig.height = 3.2}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
qqnorm(fit3_metafor, main = "metafor: studentized")
qqnorm(fit3_brma, main = "RoBMA: LOO-PIT", reps = 500)
par(op)
```

`influence()` collects DFFITS, Cook's distance, COVRATIO, leverage, and DFBETAS where the diagnostic is available.

```{r influence}
inf_brma <- influence(fit3_brma)
head(inf_brma$inf)
```

The Bayesian framework adds Pareto-k diagnostics from the `loo` object, flagging studies for which the LOO approximation is unreliable [@vehtari2024pareto].

```{r loo-plot, fig.width = 4.5, fig.height = 3.5}
plot(loo(fit3_brma))
```

## General Considerations and Reporting

`brma()` and `metafor::rma()` share an interface, but the inferences they support diverge once the fit is in hand.

**Posterior summaries replace point estimates.**
Report the posterior mean (or median) together with the credible interval, not just a single value.
A Bayesian credible interval is a probability statement about the parameter conditional on the data and the prior, which is not the same as a frequentist confidence interval.

**Bayes factors and posterior probabilities depend on the prior.**
The Bayes factor and posterior model probabilities are functions of the priors under each model, not of the data alone.
Report the priors used (or refer to the *Prior Distributions* vignette when the defaults apply) and run a prior sensitivity analysis when conclusions are close to a decision threshold.

**Inspect Pareto-k for LOO-based diagnostics.**
LOO-PIT residuals, model comparison via PSIS-LOO [@vehtari2017practical], and the Pareto-k diagnostic all rely on the same leave-one-out approximation.
Report any studies whose Pareto-k values exceed the diagnostic threshold; a small number of high-k points signals that LOO is unreliable for those studies.

## References
