---
title: "Feature Coverage"
author: "Frantisek Bartos"
date: "29th of April 2026"
output:
  rmarkdown::html_vignette:
    self_contained: yes
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Feature Coverage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)

tick    <- '<span class="yes">&#10003;</span>'
limited <- '<span class="limited">limited</span>'
no      <- '<span class="no">No</span>'
na_cell <- '<span class="na-cell">n/a</span>'

model_columns <- c(
  "brma.norm", "brma.glmm", "bselmodel", "bPET/bPEESE",
  "BMA.norm", "BMA.glmm", "RoBMA"
)

status_map <- c(
  Y = tick,
  L = limited,
  N = no,
  A = na_cell
)

feature_row <- function(topic, feature, pattern) {

  values <- strsplit(pattern, "", fixed = TRUE)[[1]]
  if (length(values) != length(model_columns)) {
    stop("Status pattern has wrong length: ", pattern, call. = FALSE)
  }

  out <- c(
    list(Topic = topic, Feature = feature),
    stats::setNames(as.list(status_map[values]), model_columns)
  )

  return(out)
}

feature_table <- function(rows) {

  rows <- as.data.frame(do.call(rbind, rows), stringsAsFactors = FALSE, check.names = FALSE)
  rows[] <- lapply(rows, function(x) {

    if (is.list(x)) {
      return(vapply(x, function(y) paste0(as.character(y), collapse = ""), character(1)))
    }
    return(as.character(x))
  })

  cat('
<style>
.feature-matrix {
  border-collapse: collapse;
  width: 100%;
  font-size: 0.84em;
}
.feature-matrix th,
.feature-matrix td {
  border: 1px solid #d0d7de;
  padding: 4px 6px;
  vertical-align: top;
}
.feature-matrix th {
  background: #f6f8fa;
}
.feature-matrix .topic-head {
  min-width: 38px;
  width: 38px;
}
.feature-matrix .feature-head {
  min-width: 25em;
}
.feature-matrix .model-head {
  height: 118px;
  min-width: 34px;
  max-width: 34px;
  width: 34px;
  padding: 0;
  vertical-align: bottom;
}
.feature-matrix .rotated-header {
  display: inline-block;
  transform: rotate(-90deg);
  transform-origin: 18px 18px;
  white-space: nowrap;
  width: 110px;
}
.feature-matrix .topic-cell {
  background: #f6f8fa;
  min-width: 38px;
  width: 38px;
  padding: 0;
  text-align: center;
  vertical-align: middle;
}
.feature-matrix .topic-cell span {
  display: inline-block;
  transform: rotate(-90deg);
  white-space: nowrap;
  font-weight: 700;
}
.feature-matrix .feature-cell {
  min-width: 25em;
}
.feature-matrix .status-cell {
  text-align: center;
  vertical-align: middle;
  white-space: nowrap;
  width: 34px;
}
.feature-matrix .yes {
  color: #1a7f37;
  font-weight: 700;
}
.feature-matrix .limited {
  color: #9a6700;
}
.feature-matrix .no,
.feature-matrix .na-cell {
  color: #6e7781;
}
</style>
')

  escape_html <- function(x) {

    x <- as.character(x)
    x <- gsub("&", "&amp;", x, fixed = TRUE)
    x <- gsub("<", "&lt;", x, fixed = TRUE)
    x <- gsub(">", "&gt;", x, fixed = TRUE)
    x <- gsub('"', "&quot;", x, fixed = TRUE)
    x <- gsub("`([^`]+)`", "<code>\\1</code>", x, perl = TRUE)
    return(x)
  }

  topic_runs <- rle(rows$Topic)
  topic_start <- cumsum(c(1L, head(topic_runs$lengths, -1L)))
  topic_span <- rep(0L, nrow(rows))
  topic_span[topic_start] <- topic_runs$lengths

  cat('<div style="overflow-x: auto;">\n')
  cat('<table class="feature-matrix">\n')
  cat("<thead><tr>")
  cat('<th class="topic-head">Topic</th>')
  cat('<th class="feature-head">Feature</th>')
  for (model_column in model_columns) {
    cat('<th class="model-head"><span class="rotated-header">', model_column, "</span></th>", sep = "")
  }
  cat("</tr></thead>\n")
  cat("<tbody>\n")
  for (i in seq_len(nrow(rows))) {
    cat("<tr>")
    if (topic_span[i] > 0L) {
      cat(
        '<td class="topic-cell" rowspan="', topic_span[i], '"><span>',
        escape_html(rows$Topic[i]),
        "</span></td>",
        sep = ""
      )
    }
    cat('<td class="feature-cell">', escape_html(rows$Feature[i]), "</td>", sep = "")
    for (model_column in model_columns) {
      cat('<td class="status-cell">', rows[[model_column]][i], "</td>", sep = "")
    }
    cat("</tr>\n")
  }
  cat("</tbody>\n")
  cat("</table>\n")
  cat("</div>\n")
}
```

This vignette is a development snapshot of the current `RoBMA` feature surface [@RoBMA].
The table is organized by substantive capability.
Rows that name broader `metafor` capabilities are included only as reference points for users moving between packages, not as requirements that `RoBMA` must mirror exactly [@metafor].

The `brma.norm` column refers to `brma()`, the normal likelihood interface with class `"brma.norm"`.
The `bPET/bPEESE` column summarizes both small-study-effect regression wrappers.
The `BMA.norm` column also covers the exported `BMA()` alias.
Green ticks mark available features.
Cells marked `limited` mean the feature exists, but with narrower model coverage or a different Bayesian implementation.

```{r feature-matrix, echo = FALSE, results = "asis"}
rows <- list(
  feature_row("Model / Structure", "Fixed- and random-effects models", "YYYYYYY"),
  feature_row("Model / Structure", "Moderation (`mods`)", "YYYYYYY"),
  feature_row("Model / Structure", "Location-scale models (`scale`)", "YYYYYYY"),
  feature_row("Model / Structure", "Multilevel models (`cluster`)", "YYYYYYY"),
  feature_row("Model / Structure", "Model averaging", "NNNNYYY"),
  feature_row("Model / Structure", "Inclusion Bayes factors", "NNNNYYY"),
  feature_row("Model / Structure", "General multivariate / covariance structures", "NNNNNNN"),
  feature_row("Model / Structure", "Robust / sandwich inference", "NNNNNNN"),

  feature_row("Priors", "Default Bayesian priors", "YYYYYYY"),
  feature_row("Priors", "Custom parameter priors", "YYYYYYY"),
  feature_row("Priors", "Informed priors", "YLYYYLY"),
  feature_row("Priors", "Prior-only inspection", "YYYYYYY"),

  feature_row("Publication Bias", "Selection models", "NNYNNNY"),
  feature_row("Publication Bias", "PET / PEESE models", "NNNYNNY"),
  feature_row("Publication Bias", "Model averaging over bias models", "NNNNNNY"),
  feature_row("Publication Bias", "Bias-adjusted summaries / predictions", "NNYYNNY"),
  feature_row("Publication Bias", "Classical bias tests / trim-fill / fail-safe N", "NNNNNNN"),

  feature_row("Prediction", "Pooled effect / heterogeneity summaries", "YYYYYYY"),
  feature_row("Prediction", "Prediction for new covariate values", "YYYYYYY"),
  feature_row("Prediction", "Posterior predictive response summaries", "YYYYYYY"),
  feature_row("Prediction", "True effects / BLUPs / random effects", "YYYYYYY"),
  feature_row("Prediction", "Estimated marginal means", "YYYYYYY"),
  feature_row("Prediction", "Fitted-value extraction with `fitted()`", "NNNNNNN"),

  feature_row("Residuals / Diagnostics", "Raw residuals", "YYYYYYY"),
  feature_row("Residuals / Diagnostics", "Pearson / standardized residuals", "YNNYYNL"),
  feature_row("Residuals / Diagnostics", "LOO-PIT / studentized residuals", "YYYYYYY"),
  feature_row("Residuals / Diagnostics", "Influence diagnostics", "YLLYYLL"),
  feature_row("Residuals / Diagnostics", "LOO / WAIC diagnostics", "YYYYYYY"),
  feature_row("Residuals / Diagnostics", "Moderator collinearity diagnostics", "YYYYYYY"),
  feature_row("Residuals / Diagnostics", "Refit leave-one-out / permutation tests", "NNNNNNN"),

  feature_row("Comparison", "LOO / WAIC comparison", "YYYYYYY"),
  feature_row("Comparison", "Bayes factors", "YYYYYYY"),
  feature_row("Comparison", "Frequentist LRT / omnibus tests", "NNNNNNN"),
  feature_row("Comparison", "AIC / BIC workflows", "NNNNNNN"),

  feature_row("Plots", "Posterior / prior plots", "YYYYYYY"),
  feature_row("Plots", "Funnel plots", "YYYYYYY"),
  feature_row("Plots", "Regression plots", "YYYYYYY"),
  feature_row("Plots", "Radial / Galbraith plots", "YLYYYLY"),
  feature_row("Plots", "MCMC diagnostic plots", "YYYYYYY"),
  feature_row("Plots", "Bias-model plots", "NNYYNNY"),
  feature_row("Plots", "Forest / profile / Baujat / GOSH / L'Abbe plots", "NNNNNNN"),

  feature_row("Extraction", "Posterior draw extraction", "YYYYYYY"),
  feature_row("Extraction", "Classical extraction generics (`vcov`, `confint`, `weights`, `simulate`)", "NNNNNNN")
)

feature_table(rows)
```

## References
