---
title: "Publication-Bias Adjustment"
author: "František Bartoš"
date: "28th of April 2026"
output:
  rmarkdown::html_vignette:
    self_contained: yes
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Publication-Bias Adjustment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---

```{r setup, include = FALSE}
source("vignette-cache.R", local = knitr::knit_global())

cached_fits <- vignette_cache(
  name     = "11-metafor-parity-publication-bias",
  objects  = c("fit_PET_brma", "fit_PEESE_brma", "fit_selmodel_brma"),
  packages = c("metafor", "metadat")
)

knitr::opts_chunk$set(
  collapse    = TRUE,
  comment     = "#>",
  eval        = vignette_cache_eval(cached_fits),
  message     = FALSE,
  warning     = FALSE,
  fig.width   = 7,
  fig.height  = 3.5,
  dev         = "png",
  fig.retina  = 3
)
if (.Platform$OS.type == "windows") {
  knitr::opts_chunk$set(dev.args = list(type = "cairo"))
}
```

```{r load-models, include = FALSE}
library("metafor")
library("RoBMA")

data("dat.lehmann2018", package = "metadat")
dat <- dat.lehmann2018

vignette_cache_load(cached_fits)
```

```{r precompute-models, include = FALSE, eval = FALSE}
library("metafor")
library("RoBMA")

data("dat.lehmann2018", package = "metadat")
dat <- dat.lehmann2018

fit_PET_brma <- bPET(
  yi = yi, vi = vi, measure = "SMD",
  data = dat, seed = 1
)

fit_PEESE_brma <- bPEESE(
  yi = yi, vi = vi, measure = "SMD",
  data = dat, seed = 1
)

fit_selmodel_brma <- bselmodel(
  yi = yi, vi = vi, measure = "SMD",
  data = dat, seed = 1
)

vignette_cache_save(cached_fits)
```

This vignette illustrates how to use `RoBMA` package [@RoBMA] for single-model publication-bias adjustment.
It walks through the ego-depletion dataset from `metadat`, fitting three single-model bias adjustments alongside their `metafor` counterparts [@metafor]: PET and PEESE [@stanley2014meta], and a weight-function selection model [@vevea1995general].
All of the models could also be combined with Bayesian model-averaging into a Robust Bayesian meta-analysis as described in another vignette (TODO: add link).

All `bPET()`, `bPEESE()`, and `bselmodel()` calls keep the default priors; see the [*Prior Distributions*](01-prior-distributions.html) vignette for the prior definitions and customization options.
For the non-bias-adjusted workflow showing more features (also applicable to the publication-bias models) see the [*Bayesian Meta-Analysis*](02-bayesian-meta-analysis.html) vignette.

## Default Bias-Adjustment Priors

The three model types use a separate prior on the bias-adjustment parameter, independent of the priors on `mu`, `tau`, and the meta-regression coefficients.

| Model         | Bias parameter | Default prior |
|:--------|:---------------|:------------------------------------------------------------------|
| `bPET()`      | `PET`   | Cauchy($0$, $1$), truncated to $[0, \infty)$ |
| `bPEESE()`    | `PEESE` | Cauchy($0$, $5 \cdot \text{UISD}_{\text{ratio}}$), truncated to $[0, \infty)$ |
| `bselmodel()` | `omega` | one-sided weight function with `steps = c(0.025)` and `alpha = c(1, 1)` |

The PET prior is invariant to the effect-size measure.
The PEESE scale is rescaled by $\text{UISD}_{\text{ratio}} = \text{UISD}_{\text{SMD}} / \text{UISD}_{\text{measure}}$ so the prior describes a comparable bias on each scale; the ratio is $1$ for `measure = "SMD"`.
The weight-function default specifies a Dirichlet prior on the relative publication probabilities of one-sided $p$-values below and above $0.025$, equivalent to a three-parameter selection model (3PSM).
Unlike the meta-analytic priors, the bias-adjustment priors are not affected by `rescale_priors`; they can be replaced via `prior_bias`, and the weight-function `steps` can be changed directly via the `steps` argument.

These priors corresponds to the default settings tested in the larger RoBMA-PSMA ensemble in @bartos2021no.
Note that the default priors are restricted in a way that the no publication bias case is on the parameter space boundary. 
This is not an issue for the RoBMA ensemble that includes models assuming absence of publication bias; however, these simple models might tend to produce slight downward bias if publication bias is absent.  

The publication bias adjustment models rely on the `effect_direction` argument.
Selection models penalize one-sided $p$-values evaluated against the expected sign of the effect, and the truncated PET and PEESE priors constrain the small-study coefficient to pull the estimate toward zero from the dominant direction.
The `effect_direction` argument tells the model which side that is.
The default `effect_direction = "detect"` infers the direction from the median of `yi` (positive median yields `"positive"`, negative median yields `"negative"`).
Set `effect_direction = "positive"` or `effect_direction = "negative"` explicitly when the dominant direction is known a priori.

The `dat.lehmann2018` dataset from `metadat` collects 81 standardized mean differences from the ego-depletion meta-analysis.
Effect sizes `yi` and sampling variances `vi` are already provided.

```{r data}
data("dat.lehmann2018", package = "metadat")
dat <- dat.lehmann2018
```

## PET

The PET model regresses effect sizes on the standard error.
With `metafor::rma()` it is fit by passing `mods = ~ sqrt(vi)`; the slope on `sqrt(vi)` is the PET coefficient.
RoBMA implements PET as a meta-regression (with a random-effects heterogeneity component) rather than the UWLS variant, for consistency with the rest of the RoBMA framework.

```{r fit-PET-metafor}
fit_PET_metafor <- metafor::rma(yi, vi, mods = ~ sqrt(vi), data = dat)
fit_PET_metafor
```

`bPET()` fits the same model with the default Cauchy prior on the (positive) PET coefficient.
`measure = "SMD"` selects the default priors for standardized mean differences.

```{r fit-PET-brma, eval = FALSE}
fit_PET_brma <- bPET(
  yi = yi, vi = vi, measure = "SMD",
  data = dat, seed = 1
)
```
```{r fit-PET-brma-summary}
summary(fit_PET_brma, include_mcmc_diagnostics = FALSE)
```

The posterior mean of `mu` is the bias-corrected average effect; the `PET` coefficient quantifies how strongly small studies pull the meta-analytic mean.
The left panel shows the prior and posterior of the `PET` coefficient with `prior = TRUE` overlaying the prior; the right panel shows the implied PET regression of effect on standard error from `plot_pet_peese()`, with the prior band overlaid.

```{r PET-posterior, fig.width = 7, fig.height = 3.2}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1), cex.axis = 0.8, cex.lab = 0.8, cex.main = 0.75)
plot(fit_PET_brma,        parameter = "PET", prior = TRUE, main = "PET coefficient", xlim = c(0, 4))
plot_pet_peese(fit_PET_brma,                  prior = TRUE, main = "PET regression")
```

The half-Cauchy prior on `PET` is wide; the posterior concentrates near $1$, broadly consistent with the metafor estimate but pulled toward zero by the prior.
The implied regression line intersects the distribution of effect sizes and standard errors.

## PEESE

The PEESE model regresses effect sizes on the sampling variance instead of the standard error.
With `metafor::rma()` it is fit by passing `mods = ~ vi`; the slope on `vi` is the PEESE coefficient.
As with PET, RoBMA's PEESE is the meta-regression version (retaining the heterogeneity component), not the UWLS variant.

```{r fit-PEESE-metafor}
fit_PEESE_metafor <- metafor::rma(yi, vi, mods = ~ vi, data = dat)
fit_PEESE_metafor
```

`bPEESE()` fits the same model with the default Cauchy prior on the (positive) PEESE coefficient, with the prior scale rescaled to the effect-size measure.

```{r fit-PEESE-brma, eval = FALSE}
fit_PEESE_brma <- bPEESE(
  yi = yi, vi = vi, measure = "SMD",
  data = dat, seed = 1
)
```

```{r fit-PEESE-brma-summary}
summary(fit_PEESE_brma, include_mcmc_diagnostics = FALSE)
```

As for PET, the left panel shows the prior and posterior of the `PEESE` coefficient and the right panel shows the implied PEESE regression of effect on sampling variance.

```{r PEESE-posterior, fig.width = 7, fig.height = 3.2}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1), cex.axis = 0.8, cex.lab = 0.8, cex.main = 0.75)
plot(fit_PEESE_brma,        parameter = "PEESE", prior = TRUE, main = "PEESE coefficient", xlim = c(0, 8))
plot_pet_peese(fit_PEESE_brma,                    prior = TRUE, main = "PEESE regression")
```

The bias-corrected effect under PEESE is also closer to zero than the random-effects estimate due to the weakly informative priors.

## Weight Function Selection Model

The selection model multiplies the likelihood of each study by a $p$-value-dependent publication probability.
With `metafor`, a one-sided three-parameter model is fit by passing the random-effects fit to `metafor::selmodel()`.

```{r fit-selmodel-metafor}
fit_rma_metafor      <- metafor::rma(yi, vi, data = dat, method = "ML")
fit_selmodel_metafor <- metafor::selmodel(fit_rma_metafor, type = "stepfun", steps = .025, decreasing = TRUE)
fit_selmodel_metafor
```

`bselmodel()` fits the same model in one call.
The default weight function uses a single one-sided cutoff at $0.025$.
Pass `steps = c(0.025, 0.5)` (or any other vector) to specify a finer step function; the prior on the omega vector is uniform across the implied intervals.
RoBMA's default Dirichlet prior on `omega` enforces a monotonically decreasing weight function (non-significant intervals cannot exceed the reference interval), which matches `metafor::selmodel(..., decreasing = TRUE)` above.

```{r fit-selmodel-brma, eval = FALSE}
fit_selmodel_brma <- bselmodel(
  yi = yi, vi = vi, measure = "SMD",
  data = dat, seed = 1
)
```
```{r fit-selmodel-brma-summary}
summary(fit_selmodel_brma, include_mcmc_diagnostics = FALSE)
```

The omega vector is reported with `omega[0, 0.025]` fixed to $1$ as the reference interval; `omega[0.025, 1]` is the publication probability of one-sided $p$-values above $0.025$ relative to the significant interval.
Values below $1$ indicate that non-significant studies are reported less often than significant ones.

The omega vector has one panel per interval (the reference $[0, 0.025]$ panel shows the spike at $1$).
`plot_weightfunction()` then visualizes the same weights as a step function over the one-sided $p$-value scale, with the prior band overlaid.

```{r selmodel-omega, fig.width = 7, fig.height = 3.2}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1), cex.axis = 0.8, cex.lab = 0.8, cex.main = 0.75)
plot(fit_selmodel_brma, parameter = "omega", prior = TRUE, xlim = c(0, 1))
```

```{r selmodel-weightfunction, fig.width = 4.5, fig.height = 3.2}
par(mar = c(4, 4, 2, 1), cex.axis = 0.8, cex.lab = 0.8, cex.main = 0.75)
plot_weightfunction(fit_selmodel_brma, prior = TRUE, main = "Weight-function")
```

The bias-corrected `mu` posterior is shifted toward zero relative to a `brma()` random-effects fit, with the magnitude of the shift driven by the gap between the prior and posterior of `omega[0.025, 1]`.

## Asymmetric Sampling Funnels

`funnel()` extends the standard funnel plot by drawing the *full sampling distribution implied by the fitted model*, including the publication-bias adjustment.
With a no-bias `brma()` fit the funnel is symmetric; with a bias-adjustment model the funnel becomes asymmetric in the direction the model identifies as suppressed.
Comparing the three fits side by side shows how each adjustment encodes a different mechanism.

```{r funnel-plots, fig.width = 7, fig.height = 3.2}
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1), cex.axis = 0.8, cex.lab = 0.8, cex.main = 0.75)
funnel(fit_PET_brma,      main = "PET",      xlim = c(-2, 2.5), ylim = c(0, 0.8))
funnel(fit_PEESE_brma,    main = "PEESE",    xlim = c(-2, 2.5), ylim = c(0, 0.8))
funnel(fit_selmodel_brma, main = "Sel. Mod.", xlim = c(-2, 2.5), ylim = c(0, 0.8))
```

PET tilts the funnel by adding a linear effect of the standard error, so the centerline shifts away from zero in the direction of the dominant effect as precision drops.
PEESE produces the same kind of shift but curved (quadratic in standard error), so the displacement grows faster for imprecise studies and the funnel walls bow outward.
The selection-model funnel keeps the centerline on the random-effects `mu`, but the implied sampling density on each side of the $p < 0.025$ contour differs: the non-significant region is down-weighted by `omega[0.025, 1]`, so non-significant studies are predicted to appear less often than significant ones at the same precision.

## Other Inference Helpers

`bPET()`, `bPEESE()`, and `bselmodel()` accept the same `mods`, `scale`, and `cluster` arguments as `brma()`, so meta-regression, location-scale, and multilevel publication-bias models are directly available. The fits are also `brma` objects, so all summaries, plots, predictions, and diagnostics examined in the [*Bayesian Meta-Analysis*](02-bayesian-meta-analysis.html) vignette work the same way here.

To average across publication-bias models rather than commit to a single one, use `RoBMA()`; see the [*Introduction to RoBMA*](00-introduction.html) vignette for an overview of the model-averaged workflow and the available `model_type` presets.

## References
