---
title: "Multilevel Robust Bayesian Model-Averaged Meta-Regression"
author: "František Bartoš"
date: "17th of December 2025 (updated: 29th of April 2026)"
output:
  rmarkdown::html_vignette:
    self_contained: yes
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Multilevel Robust Bayesian Model-Averaged Meta-Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---

```{r setup, include = FALSE}
source("vignette-cache.R", local = knitr::knit_global())

cached_fits <- vignette_cache(
  name     = "32-robma-multilevel",
  objects  = c("fit_reg", "zfit_reg")
)

knitr::opts_chunk$set(
  collapse    = TRUE,
  comment     = "#>",
  eval        = vignette_cache_eval(cached_fits),
  message     = FALSE,
  warning     = FALSE,
  fig.width   = 7,
  fig.height  = 3.5,
  dev         = "png",
  fig.retina  = 3
)
if (.Platform$OS.type == "windows") {
  knitr::opts_chunk$set(dev.args = list(type = "cairo"))
}
```

```{r load-models, include = FALSE}
library("RoBMA")

data("Havrankova2025", package = "RoBMA")

unit_information_sd_361 <- estimate_unit_information_sd(
  sei = Havrankova2025$se,
  ni  = Havrankova2025$N
)
prior_scale_361 <- unit_information_sd_361 * 0.5

vignette_cache_load(cached_fits)
```

```{r precompute-models, include = FALSE, eval = FALSE}
library("RoBMA")

data("Havrankova2025", package = "RoBMA")

unit_information_sd_361 <- estimate_unit_information_sd(
  sei = Havrankova2025$se,
  ni  = Havrankova2025$N
)
prior_scale_361 <- unit_information_sd_361 * 0.5

prior_effect_361        <- prior("normal",   list(mean = 0, sd = prior_scale_361))
prior_heterogeneity_361 <- prior("invgamma", list(shape = 1, scale = 0.15 * prior_scale_361))
prior_mods_361          <- list(
  facing_customer = prior_factor("mnormal", list(mean = 0, sd = 0.25 * prior_scale_361), contrast = "meandif")
)
prior_bias_361          <- list(
  prior_weightfunction("two-sided", c(0.05),             wf_cumulative(c(1, 1)),       1/12),
  prior_weightfunction("two-sided", c(0.05, 0.10),       wf_cumulative(c(1, 1, 1)),    1/12),
  prior_weightfunction("one-sided", c(0.05),             wf_cumulative(c(1, 1)),       1/12),
  prior_weightfunction("one-sided", c(0.025, 0.05),      wf_cumulative(c(1, 1, 1)),    1/12),
  prior_weightfunction("one-sided", c(0.05, 0.5),        wf_cumulative(c(1, 1, 1)),    1/12),
  prior_weightfunction("one-sided", c(0.025, 0.05, 0.5), wf_cumulative(c(1, 1, 1, 1)), 1/12),
  prior_PET("Cauchy", list(0, 1), list(0, Inf), 1/4),
  prior_PEESE("Cauchy", list(0, 5 / prior_scale_361), list(0, Inf), 1/4)
)

fit_reg <- RoBMA(
  yi = y, sei = se, ni = N, measure = "GEN",
  mods = ~ facing_customer,
  cluster = study_id,
  prior_unit_information_sd = unit_information_sd_361,
  prior_effect              = prior_effect_361,
  prior_heterogeneity       = prior_heterogeneity_361,
  prior_mods                = prior_mods_361,
  prior_bias                = prior_bias_361,
  sample = 20000, burnin = 10000, adapt = 10000,
  thin = 5, parallel = TRUE, seed = 1,
  data = Havrankova2025
)

zfit_reg <- as_zcurve(fit_reg)

vignette_cache_save(cached_fits)
```

**This vignette accompanies the manuscript [Robust Bayesian Multilevel Meta-Analysis: Adjusting for Publication Bias in the Presence of Dependent Effect Sizes](https://doi.org/10.31234/osf.io/9tgp2_v1) preprinted at *PsyArXiv* [@bartos2025robust].**

This vignette reproduces the second example from the manuscript, re-analyzing data from @havrankova2025beauty on the relationship between beauty and professional success (1,159 effect sizes from 67 studies).
We investigate whether the effect depends on the type of customer contact ("no", "some", or "direct").
For the basic multilevel syntax see the [*Multilevel Meta-Analysis*](10-metafor-parity-multilevel.html) vignette.

## Rescaling Priors for Non-Standardized Effect Sizes

The effect sizes in this dataset are not standardized: they are percent increases in earnings with one-standard deviation increase in beauty.
Therefore, the prior distributions need to be defined on this outcome scale.
Overly narrow priors would make it difficult to find evidence for the effect because both models would predict effects very close to zero.
Overly wide priors would bias the results towards the null hypothesis because alternative models would predict unrealistically large effects.

We rescale the RoBMA 3.6.1 priors by matching the default prior scale to the unit information scale of the data [@rover2021weakly; @mulder2024bayesian].
The unit information scale can be approximated by the standard error of a fixed-effect model multiplied by the square root of the total sample size.
The old default effect prior was specified on Cohen's $d$ with standard deviation one, which corresponds to half the unit information scale under the package's approximate Cohen's $d$ to Fisher's $z$ scaling.
We therefore multiply the unit information scale by 0.5.[^scaling]

[^scaling]: This reproduces the RoBMA 3.6.1 vignette results. It does not match the current default UISD scaling conventions exactly; in particular, the current default moderator and PEESE scales differ and are therefore specified below.

```{r scale-priors}
library("RoBMA")

data("Havrankova2025", package = "RoBMA")

unit_information_sd_361 <- estimate_unit_information_sd(
  sei = Havrankova2025$se,
  ni  = Havrankova2025$N
)
prior_scale_361 <- unit_information_sd_361 * 0.5
prior_scale_361
```

The `Havrankova2025` data already contain study-level estimates and standard errors.
We can therefore use the current unified input interface directly with `yi = y`, `sei = se`, and `ni = N`.
The exported `estimate_unit_information_sd()` helper computes the unit information scale from standard errors and sample sizes.

The old prior specification is made explicit below:

```{r prior-specification}
prior_effect_361        <- prior("normal",   list(mean = 0, sd = prior_scale_361))
prior_heterogeneity_361 <- prior("invgamma", list(shape = 1, scale = 0.15 * prior_scale_361))
prior_mods_361          <- list(
  facing_customer = prior_factor("mnormal", list(mean = 0, sd = 0.25 * prior_scale_361), contrast = "meandif")
)
prior_bias_361          <- list(
  prior_weightfunction("two-sided", c(0.05),             wf_cumulative(c(1, 1)),       1/12),
  prior_weightfunction("two-sided", c(0.05, 0.10),       wf_cumulative(c(1, 1, 1)),    1/12),
  prior_weightfunction("one-sided", c(0.05),             wf_cumulative(c(1, 1)),       1/12),
  prior_weightfunction("one-sided", c(0.025, 0.05),      wf_cumulative(c(1, 1, 1)),    1/12),
  prior_weightfunction("one-sided", c(0.05, 0.5),        wf_cumulative(c(1, 1, 1)),    1/12),
  prior_weightfunction("one-sided", c(0.025, 0.05, 0.5), wf_cumulative(c(1, 1, 1, 1)), 1/12),
  prior_PET("Cauchy", list(0, 1), list(0, Inf), 1/4),
  prior_PEESE("Cauchy", list(0, 5 / prior_scale_361), list(0, Inf), 1/4)
)
```

Together with `prior_unit_information_sd = unit_information_sd_361`, these priors reproduce the RoBMA 3.6.1 defaults on the non-standardized outcome scale.
The effect prior matches what the current defaults would imply after setting the UISD, but it is specified explicitly for consistency with the old inverse-gamma heterogeneity prior, moderator scale, and PEESE scale.

We fit the robust Bayesian multilevel meta-regression with the current unified input interface.
The model uses the `y` and `se` columns as effect size and standard error inputs, the `facing_customer` moderator, and `study_id` as the clustering variable.

```{r fit-reg, eval = FALSE}
fit_reg <- RoBMA(
  yi = y, sei = se, ni = N, measure = "GEN",
  mods = ~ facing_customer,
  cluster = study_id,
  prior_unit_information_sd = unit_information_sd_361,
  prior_effect              = prior_effect_361,
  prior_heterogeneity       = prior_heterogeneity_361,
  prior_mods                = prior_mods_361,
  prior_bias                = prior_bias_361,
  sample = 20000, burnin = 10000, adapt = 10000,
  thin = 5, parallel = TRUE, seed = 1,
  data = Havrankova2025
)
```

The meta-regression can take more than an hour to run with parallel processing enabled.

## Interpreting the Results

The `summary()` function reports inclusion Bayes factors and model parameters.
The effect-size estimates are obtained with dedicated helpers: `pooled_effect()` for the average effect and `marginal_means()` for the moderator levels.

```{r summary}
summary(fit_reg, include_mcmc_diagnostics = FALSE)
```

Using the rescaled 3.6.1 prior distributions, the multilevel RoBMA meta-regression finds extreme evidence for the presence of the average effect, moderation by the degree of consumer contact, between-study heterogeneity, and publication bias (all BFs > 10^6).

```{r pooled-effect}
pooled_effect(fit_reg)
```

The model-averaged average effect estimate is accompanied by considerable overall heterogeneity and cluster-level allocation close to 1, indicating a high degree of similarity of estimates from the same study.
MCMC diagnostics are included in `summary()` by default, and chain-level checks are available via `plot_diagnostic_trace()`, `plot_diagnostic_density()`, and `plot_diagnostic_autocorrelation()`.

The average effect is moderated by the degree of consumer contact.
Estimated marginal means give the effect at each level of the moderator.

```{r marginal-means}
mm_reg <- marginal_means(fit_reg)
summary(mm_reg)
```

The estimated marginal means reveal moderate evidence for the absence of the effect for jobs with no consumer contact and extreme evidence for jobs with some or direct consumer contact.

## Visual Fit Assessment

We visualize model fit using the meta-analytic z-curve plot [@bartos2025zcurve], also see the [*Z-Curve Publication Bias Diagnostics*](35-zplot.html) vignette for more details.
The z-curve plot shows the distribution of the observed $z$-statistics, computed as the effect size divided by the standard error, with dotted red horizontal lines highlighting the typical steps on which publication bias operates.
These are $z = \pm 1.65$ and $z = \pm 1.96$, corresponding to marginally significant and statistically significant $p$-values, and $z = 0$, corresponding to selection on the direction of the effect.

```{r as-zcurve, eval = FALSE}
zfit_reg <- as_zcurve(fit_reg)
```

```{r zcurve, fig.width = 6, fig.height = 4}
par(mar = c(4, 4, 0, 0))
plot(zfit_reg, by.hist = 0.25, plot_extrapolation = FALSE, from = -4, to = 8)
```

We can notice clear discontinuities corresponding to selection on the direction of the effect and marginal significance.
The posterior predictive density from the multilevel RoBMA meta-regression indicates that the model approximates the observed data reasonably well, capturing the discontinuities and the proportion of statistically significant results.
This supports the model fit while adjusting for publication bias.

## References
