---
title: "Feature Parity with metafor"
author: "Frantisek Bartos"
date: "27th of April 2026"
output:
  rmarkdown::html_vignette:
    self_contained: yes
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Feature Parity with metafor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---

```{r setup, include = FALSE}
cached_models <- file.path(
  "../models/MetaforParity",
  c(
    "fit1_brma.RDS",
    "fit2_brma.RDS",
    "fit3_brma.RDS",
    "fit_bma.RDS",
    "fit_robma_pp.RDS"
  )
)

has_suggested <- requireNamespace("metafor", quietly = TRUE) &&
  requireNamespace("metadat", quietly = TRUE)

is_check <- ("CheckExEnv" %in% search()) ||
  any(c("_R_CHECK_TIMINGS_", "_R_CHECK_LICENSE_") %in% names(Sys.getenv())) ||
  !has_suggested ||
  !all(file.exists(cached_models))

knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  eval       = !is_check,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 3.5,
  dev        = "png"
)
if (.Platform$OS.type == "windows") {
  knitr::opts_chunk$set(dev.args = list(type = "cairo"))
}
```

```{r load-models, include = FALSE}
library("metafor")
library("RoBMA")

data("dat.bcg", package = "metadat")

dat <- metafor::escalc(
  measure = "RR",
  ai      = tpos,
  bi      = tneg,
  ci      = cpos,
  di      = cneg,
  data    = dat.bcg
)

fit1_metafor <- metafor::rma(yi, vi, data = dat, method = "REML")
fit2_metafor <- metafor::rma(yi, vi, mods = ~ ablat, data = dat)
fit3_metafor <- metafor::rma(yi, vi, mods = ~ ablat + alloc, data = dat)

fit1_brma    <- readRDS(file = "../models/MetaforParity/fit1_brma.RDS")
fit2_brma    <- readRDS(file = "../models/MetaforParity/fit2_brma.RDS")
fit3_brma    <- readRDS(file = "../models/MetaforParity/fit3_brma.RDS")
fit_bma      <- readRDS(file = "../models/MetaforParity/fit_bma.RDS")
fit_robma_pp <- readRDS(file = "../models/MetaforParity/fit_robma_pp.RDS")
```

```{r precompute-models, include = FALSE, eval = FALSE}
library("metafor")
library("RoBMA")

data("dat.bcg", package = "metadat")

dat <- metafor::escalc(
  measure = "RR",
  ai      = tpos,
  bi      = tneg,
  ci      = cpos,
  di      = cneg,
  data    = dat.bcg
)

dir.create("../models/MetaforParity", showWarnings = FALSE)

fit1_brma <- brma(
  yi       = yi,
  vi       = vi,
  data     = dat,
  measure  = "RR",
  chains   = 2,
  sample   = 1000,
  burnin   = 500,
  adapt    = 500,
  seed     = 1,
  silent   = TRUE,
  parallel = FALSE
)
fit1_brma <- add_loo(fit1_brma)
fit1_brma <- add_marglik(fit1_brma)

fit2_brma <- brma(
  yi       = yi,
  vi       = vi,
  mods     = ~ ablat,
  data     = dat,
  measure  = "RR",
  chains   = 2,
  sample   = 1000,
  burnin   = 500,
  adapt    = 500,
  seed     = 2,
  silent   = TRUE,
  parallel = FALSE
)
fit2_brma <- add_loo(fit2_brma)
fit2_brma <- add_marglik(fit2_brma)

fit3_brma <- brma(
  yi       = yi,
  vi       = vi,
  mods     = ~ ablat + alloc,
  data     = dat,
  measure  = "RR",
  chains   = 2,
  sample   = 1000,
  burnin   = 500,
  adapt    = 500,
  seed     = 3,
  silent   = TRUE,
  parallel = FALSE
)
fit3_brma <- add_loo(fit3_brma)
fit3_brma <- add_marglik(fit3_brma)

fit_bma <- BMA(
  yi       = yi,
  vi       = vi,
  data     = dat,
  measure  = "RR",
  chains   = 2,
  sample   = 1000,
  burnin   = 500,
  adapt    = 500,
  seed     = 4,
  silent   = TRUE,
  parallel = FALSE
)

fit_robma_pp <- RoBMA(
  yi         = yi,
  vi         = vi,
  data       = dat,
  measure    = "RR",
  model_type = "PP",
  chains     = 2,
  sample     = 1000,
  burnin     = 500,
  adapt      = 500,
  seed       = 5,
  silent     = TRUE,
  parallel   = FALSE
)

saveRDS(fit1_brma,    file = "../models/MetaforParity/fit1_brma.RDS",    compress = "xz")
saveRDS(fit2_brma,    file = "../models/MetaforParity/fit2_brma.RDS",    compress = "xz")
saveRDS(fit3_brma,    file = "../models/MetaforParity/fit3_brma.RDS",    compress = "xz")
saveRDS(fit_bma,      file = "../models/MetaforParity/fit_bma.RDS",      compress = "xz")
saveRDS(fit_robma_pp, file = "../models/MetaforParity/fit_robma_pp.RDS", compress = "xz")
```

This vignette starts from the BCG vaccine examples in `metafor` [@metafor] and fits matching Bayesian models with `brma()` [@RoBMA].
All `brma()` calls keep the default priors; see `vignette("PriorDistributions", package = "RoBMA")` for the prior definitions and practical guidance.
The outcome is the log risk ratio (`measure = "RR"`).

## Data and Models

The `metafor` workflow computes effect sizes with `escalc()` and then fits random-effects and mixed-effects models with `rma()`.

```{r metafor-models}
dat <- metafor::escalc(
  measure = "RR",
  ai      = tpos,
  bi      = tneg,
  ci      = cpos,
  di      = cneg,
  data    = dat.bcg
)

fit1_metafor <- metafor::rma(yi, vi, data = dat, method = "REML")
fit2_metafor <- metafor::rma(yi, vi, mods = ~ ablat, data = dat)
fit3_metafor <- metafor::rma(yi, vi, mods = ~ ablat + alloc, data = dat)
```

The corresponding Bayesian models use the same effect-size columns and moderator formulas.
`measure = "RR"` tells `brma()` to use the default priors for log risk ratios.

```{r brma-model-code, eval = FALSE}
fit1_brma <- brma(yi = yi, vi = vi, data = dat, measure = "RR", seed = 1)
fit2_brma <- brma(yi = yi, vi = vi, mods = ~ ablat, data = dat, measure = "RR", seed = 2)
fit3_brma <- brma(yi = yi, vi = vi, mods = ~ ablat + alloc, data = dat, measure = "RR", seed = 3)
```

`summary()` gives posterior means, credible intervals, and MCMC diagnostics.
For meta-regression, the intercept and moderator terms appear in a separate meta-regression table.

```{r summaries}
summary(fit1_brma, include_MCMC_diagnostics = FALSE)
summary(fit2_brma, include_MCMC_diagnostics = FALSE)
summary(fit3_brma, include_MCMC_diagnostics = FALSE)
```

`summary_heterogeneity()` reports absolute and relative heterogeneity summaries.
This is the Bayesian analogue of inspecting `tau^2`, `I^2`, and related quantities in `metafor`.

```{r heterogeneity}
summary_heterogeneity(fit1_brma)
summary_heterogeneity(fit3_brma)
```

Estimated marginal means are TBA as a direct `emmeans` method.
For now, use `predict()` on a grid of moderator values; the dummy `yi` and `vi` columns are only needed for input validation and are not used by `type = "terms"`.

```{r marginal-means}
emm_alloc <- expand.grid(
  ablat = mean(dat$ablat),
  alloc = sort(unique(dat$alloc))
)
emm_alloc$yi <- 0
emm_alloc$vi <- 1

predict(fit3_brma, newdata = emm_alloc, type = "terms")
```

## Bayes Factors

Unlike `metafor`, `brma()` models can be compared by marginal likelihoods and Bayes factors after `add_marglik()`.
These comparisons answer how much the observed data changed the relative predictive support for one model over another.

```{r bayes-factors}
bf(fit2_brma, fit1_brma)
bf(fit3_brma, fit2_brma)

post_prob(
  fit1_brma,
  fit2_brma,
  fit3_brma,
  model_names = c("intercept", "ablat", "ablat + alloc")
)
```

## Regression Plots

`regplot()` creates the familiar bubble plot.
For a continuous moderator, the `brma()` output is intentionally close to `metafor::regplot()`.

```{r regplot-continuous, fig.width = 7, fig.height = 3.2}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
metafor::regplot(fit2_metafor, main = "metafor")
regplot(fit2_brma, main = "RoBMA")
par(op)
```

For categorical moderators, `metafor` plots individual dummy columns.
`RoBMA::regplot()` can plot the factor as a factor, while still accounting for the other moderator in the fitted model.

```{r regplot-categorical, fig.width = 7, fig.height = 3.2}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
metafor::regplot(fit3_metafor, mod = "allocrandom", main = "metafor: random dummy")
regplot(fit3_brma, mod = "alloc", main = "RoBMA: alloc factor")
par(op)
```

## Funnel and Residuals

`funnel()` shows observed effect sizes against the fitted sampling distribution for intercept-only models.
For meta-regression models, `RoBMA::funnel()` automatically switches to a residual funnel.

```{r funnel-plot, fig.width = 7, fig.height = 3.2}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
metafor::funnel(fit1_metafor, main = "metafor")
funnel(fit1_brma, main = "RoBMA")
par(op)
```

Residual helpers follow the usual `residuals()`, `rstandard()`, and `rstudent()` workflow.
`rstudent()` uses LOO-PIT residuals, which propagate posterior uncertainty and leverage through leave-one-out prediction.

```{r residuals}
head(residuals(fit2_brma))
head(rstandard(fit2_brma))
head(rstudent(fit2_brma))
```

## QQ and Radial Plots

`qqnorm()` checks standardized residuals against a standard normal reference.
The `RoBMA` method defaults to LOO-PIT residuals, so `add_loo()` must be run first.

```{r qqplot, fig.width = 7, fig.height = 3.2}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
metafor::qqnorm(fit1_metafor, main = "metafor")
qqnorm(fit1_brma, main = "RoBMA", reps = 500)
par(op)
```

`radial()` creates a Galbraith radial plot for intercept-only models.
Use `transf = exp` when you want the outer labels on the risk-ratio scale.

```{r radial-plot, fig.width = 7, fig.height = 3.2}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
metafor::radial(fit1_metafor, main = "metafor")
radial(fit1_brma, main = "RoBMA")
par(op)
```

## Influence and Diagnostics

`influence()` collects LOO-PIT residuals, DFFITS, Cook's distance, COVRATIO, leverage, and DFBETAS where the diagnostic is available.
The individual helper functions can be called directly when a focused diagnostic is needed.

```{r influence-table}
inf_brma <- influence(fit1_brma)
head(inf_brma$inf)

diagnostics_brma <- data.frame(
  rstudent = rstudent(fit1_brma)$z,
  dffits   = dffits(fit1_brma),
  cook.d   = cooks.distance(fit1_brma),
  cov.r    = covratio(fit1_brma),
  hat      = hatvalues(fit1_brma)
)
head(diagnostics_brma)
head(dfbetas(fit1_brma))
```

The same quantities can be plotted directly from the returned values.
Here Cook's distance is shown side by side for the intercept-only model.

```{r influence-plot, fig.width = 7, fig.height = 3.2}
inf_metafor <- influence(fit1_metafor)

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(inf_metafor$inf$cook.d, type = "h", xlab = "Study", ylab = "Cook's distance", main = "metafor")
plot(inf_brma$inf$cook.d, type = "h", xlab = "Study", ylab = "Cook's distance", main = "RoBMA")
par(op)
```

## LOO

`metafor` commonly uses likelihood information criteria such as AIC.
`RoBMA` adds PSIS-LOO for Bayesian expected out-of-sample predictive performance, with Pareto-k diagnostics available from the `loo` object.

```{r loo}
AIC(fit1_metafor, fit2_metafor, fit3_metafor)

loo(fit1_brma)
loo_compare(fit1_brma, fit2_brma, fit3_brma)
```

```{r loo-plot, fig.width = 4.5, fig.height = 3.5}
plot(loo(fit1_brma))
```

## Beyond metafor

`BMA()` averages over models with and without an effect and heterogeneity.
The summary reports posterior inclusion probabilities and inclusion Bayes factors, which are not part of the standard `metafor` workflow.

```{r bma-summary}
summary(fit_bma, include_MCMC_diagnostics = FALSE)
```

`RoBMA()` extends this idea to publication-bias adjustment.
The cached example uses the PET-PEESE publication-bias ensemble; the full default `RoBMA()` call uses the PSMA ensemble, which also averages over weight-function selection models.

```{r robma-summary}
summary(fit_robma_pp, include_MCMC_diagnostics = FALSE)
```

```{r full-robma-code, eval = FALSE}
fit_robma <- RoBMA(yi = yi, vi = vi, data = dat, measure = "RR", seed = 4)
```

`metafor::selmodel()` can fit a specified selection model.
`RoBMA()` instead lets the data update a model-averaged ensemble over no-bias, PET, PEESE, and selection-model components, yielding posterior inclusion probabilities and Bayes factors for publication bias.

## References
