---
title: "Prior Distributions"
author: "František Bartoš"
date: "27th of April 2026"
output:
  rmarkdown::html_vignette:
    self_contained: yes
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Prior Distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width = 6,
  fig.height = 4
)

library(RoBMA)

has_bcg <- requireNamespace("metadat", quietly = TRUE) &&
  requireNamespace("metafor", quietly = TRUE)
```

Prior distributions form an important part of every Bayesian analysis.
They describe which parameter values are plausible before seeing the data.
In meta-analysis this matters because many datasets contain only a small or moderate number of studies.
Reasonable priors help stabilizing estimates where the data are often weak.
For estimation, weakly informative priors are often useful: they rule out extreme values and regularize estimates.

Prior distributions matter even more for Bayesian hypothesis testing and Bayesian model averaging.
A Bayes factor quantifies how well different hypotheses predicted the data.
If the alternative hypothesis uses a very wide prior, it spreads probability across many implausible values and can make ordinary data look more compatible with the null hypothesis.
In model averaging, the same issue can shrink model-averaged estimates strongly toward the null because the alternative model receives little posterior probability.
The default prior distributions in `RoBMA` are therefore not intended to be "flat".
They are weakly informative priors that provide reasonable shrinkage while still being usable for hypothesis testing and model averaging.
This discussion follows the prior-distribution guidance in @bartos2025bayesian.

## Four Ways to Set Priors

`RoBMA` offers four practical ways to specify prior distributions.

| Approach | Use when | Main arguments |
|:--|:--|:--|
| Default priors | You want a standard, defensible starting point | `measure`, `rescale_priors` (`ni` for non-standardized measures)  |
| Informed priors | You want empirical prior distributions | `prior_informed_field`, `prior_informed_subfield`, or `prior_informed()` |
| Custom priors | You know exactly what prior distributions you want | `prior_effect`, `prior_heterogeneity`, `prior_mods`, `prior_scale` |
| Publication-bias priors | You want to average across bias-adjustment models | `model_type`, `prior_bias`, `prior_bias_null` |

The first three options describe priors for the effect, heterogeneity, and meta-regression terms.
Publication-bias priors are treated separately.

The examples below use `only_priors = TRUE`.
This asks the package to set up the prior distributions and then stop before fitting the model.
Remove `only_priors = TRUE` when you want to run the analysis.

## Default Priors

The easiest way to use the defaults is to tell `RoBMA` what effect-size measure is being analyzed.
The package then specifies prior distributions on the matching scale.
We start with `brma()` because it fits one meta-analytic model and makes the priors easiest to see.

For example, the BCG vaccine dataset from the `metadat` package contains two-by-two tables.
We first compute log risk ratios and their standard errors, then ask `brma()` to show the default priors for a simple random-effects meta-analysis.

```{r bcg-data, eval = has_bcg}
data("dat.bcg", package = "metadat")

bcg <- metafor::escalc(
  measure = "RR",
  ai      = tpos,
  bi      = tneg,
  ci      = cpos,
  di      = cneg,
  data    = dat.bcg
)
```

```{r bcg-default-priors, eval = has_bcg}
bcg_priors <- brma(
  yi          = yi,
  sei         = sqrt(vi),
  data        = bcg,
  measure     = "RR",
  only_priors = TRUE
)

plot_prior(bcg_priors, parameter = "mu")
plot_prior(bcg_priors, parameter = "tau")
print_prior(bcg_priors, parameter = "mu")
print_prior(bcg_priors, parameter = "tau")
```

The first plot shows the prior for the average effect, $\mu$.
The second plot shows the prior for the between-study heterogeneity, $\tau$.
Because `brma()` fits one model, each included parameter has one prior distribution.

For `measure = "RR"`, the effect size is a log risk ratio.
The package uses a default prior of `Normal(0, 1)` for the effect and a positive `Normal(0, 0.5)` prior for heterogeneity.
These numbers come from the unit information standard deviation described in the next section.

## Unit Information Standard Deviation

The unit information standard deviation, or UISD, is a way to place priors on a scale that matches the effect-size measure.
It is useful because a standardized mean difference, a log odds ratio, and a Fisher's z transformed correlation do not have the same numerical scale.
A value of 1 can be large for one measure and moderate for another.

`RoBMA` follows the UISD idea discussed by @rover2021weakly.
For common effect-size measures, the UISD is known in advance.
For other measures, such as regression coefficients on a custom scale, the package can estimate it from the standard errors and sample sizes.

| Effect-size measure | `measure` | UISD used by default |
|:--|:--|:--|
| Standardized mean difference | `"SMD"` | `sqrt(2)` |
| Fisher's z transformed correlation | `"ZCOR"` | `1` |
| Log risk ratio | `"RR"` | `sqrt(4)` |
| Log odds ratio | `"OR"` | `sqrt(4)` |
| Log hazard ratio | `"HR"` | `sqrt(4)` |
| Log incidence rate ratio | `"IRR"` | `sqrt(4)` |
| Generic or custom scale | `"GEN"` | estimated from `sei` and `ni`, or set manually |

The defaults then use fractions of the UISD to specify the default scale of prior distributions.

| Parameter | Meaning | Default prior |
|:--|:--|:--|
| `mu` | average effect | `Normal(0, 0.5 * UISD)` |
| `tau` | heterogeneity | positive `Normal(0, 0.25 * UISD)` |
| `mods` | meta-regression coefficients | `Normal(0, 0.25 * UISD)` |
| `scale` | heterogeneity-regression coefficients | `Normal(0, 0.5)` |
| `rho` | heterogeneity allocation in multilevel models | `Beta(1, 1)` |

The `scale` regression coefficients do not use the UISD.
They describe changes in log heterogeneity, so their scale is different from effect-size priors.
The notation for meta-regression and multilevel parameters follows the robust Bayesian meta-regression and multilevel RoBMA papers [@bartos2023robust; @bartos2025robust].

You can make the default priors wider or narrower with `rescale_priors`.
Values larger than 1 make priors wider.
Values smaller than 1 make priors narrower.

```{r bcg-rescale, eval = has_bcg}
bcg_wider_priors <- brma(
  yi             = yi,
  sei            = sqrt(vi),
  data           = bcg,
  measure        = "RR",
  rescale_priors = 2,
  only_priors    = TRUE
)

print_prior(bcg_wider_priors, parameter = "mu")
print_prior(bcg_wider_priors, parameter = "tau")
```

The effect prior changed from `Normal(0, 1)` to `Normal(0, 2)`, and the heterogeneity prior changed from positive `Normal(0, 0.5)` to positive `Normal(0, 1)`.

## Correlations and Regression Coefficients

Some datasets report correlations.
Correlations should be analyzed on Fisher's z scale, not on the raw correlation scale.
In `RoBMA`, this corresponds to `measure = "ZCOR"`.

The `Kroupova2021` dataset, collected by Kroupova, Havranek, and Irsova, contains many estimates of the relationship between student employment and educational outcomes.
Here we use it to show how the package sets priors for meta-regression terms.
We first use `metafor::escalc()` to transform the reported correlations to Fisher's z and compute their sampling variances.

```{r Kroupova2021-data}
data("Kroupova2021", package = "RoBMA")

Kroupova2021 <- metafor::escalc(
  measure = "ZCOR",
  ri      = r,
  ni      = sample_size,
  data    = Kroupova2021
)
```

```{r Kroupova2021-priors}
Kroupova2021_priors <- brma(
  yi                             = yi,
  vi                             = vi,
  data                           = Kroupova2021,
  measure                        = "ZCOR",
  mods                           = ~ education_outcome + location,
  set_contrast_factor_predictors = "meandif",
  only_priors                    = TRUE
)

print_prior(Kroupova2021_priors, parameter = "mu")
print_prior(Kroupova2021_priors, parameter_mods = "education_outcome")
print_prior(Kroupova2021_priors, parameter = "tau")
```

The intercept prior is the prior for the covariate-adjusted average effect.
The moderator priors describe how much the effect may change across education outcomes or locations.
Here we set categorical predictors to use mean-difference contrasts, so the coefficients are differences from the grand mean rather than differences from one arbitrary reference category.
By default, `plot_prior()` shows meta-regression priors on the standardized coefficient scale, because this is the scale on which the default priors are specified.
For continuous moderators, use `standardized_coefficients = FALSE` to show the prior per one original unit of the moderator.

```{r Kroupova2021-plot-priors}
plot_prior(Kroupova2021_priors, parameter = "mu")
plot_prior(Kroupova2021_priors, parameter_mods = "education_outcome")
```

If the default Fisher's z prior scale is not the scale you want, set it directly.

```{r Kroupova2021-manual-uisd}
Kroupova2021_manual_priors <- brma(
  yi                             = yi,
  vi                             = vi,
  data                           = Kroupova2021,
  measure                        = "ZCOR",
  mods                           = ~ education_outcome + location,
  set_contrast_factor_predictors = "meandif",
  prior_unit_information_sd      = 0.5,
  only_priors                    = TRUE
)

print_prior(Kroupova2021_manual_priors, parameter = "mu")
print_prior(Kroupova2021_manual_priors, parameter_mods = "education_outcome")
```

Here the effect and moderator priors are narrower because `prior_unit_information_sd = 0.5` says that plausible effects should be placed on a smaller scale.

## Informed Priors

Default priors are meant to work broadly.
Informed priors use previous meta-analyses to define what effects and heterogeneity values have looked like in a particular research area.

For medical meta-analyses, `RoBMA` can use empirical priors based on the Cochrane Database of Systematic Reviews.
Continuous outcomes are described by @bartos2021bayesian, and binary or time-to-event outcomes by @bartos2023empirical.
The shortcut is `prior_informed_field = "medicine"`.
If no subfield is given, the package uses the full Cochrane database.

```{r bcg-informed, eval = has_bcg}
bcg_informed_priors <- brma(
  yi                     = yi,
  vi                     = vi,
  data                   = bcg,
  measure                = "RR",
  prior_informed_field    = "medicine",
  prior_informed_subfield = "Cochrane",
  only_priors            = TRUE
)

print_prior(bcg_informed_priors, parameter = "mu")
print_prior(bcg_informed_priors, parameter = "tau")
```

The package also re-exports `prior_informed()` from `BayesTools`.
This can be used for older informed priors from psychology, such as the informed prior distribution for the effect size from @gronau2017bayesian or the heterogeneity prior from @erp2017estimates.
These are best treated as custom priors and passed directly to the relevant arguments.

```{r psychology-informed}
psychology_priors <- brma(
  yi                  = yi,
  vi                  = vi,
  data                = Kroupova2021,
  measure             = "ZCOR",
  prior_effect        = prior_informed("Oosterwijk"),
  prior_heterogeneity = prior_informed("van Erp", parameter = "heterogeneity"),
  only_priors         = TRUE
)

print_prior(psychology_priors, parameter = "mu")
print_prior(psychology_priors, parameter = "tau")
```

Informed priors are helpful when the previous literature is relevant to the current analysis.
They should not be used only because they produce a preferred conclusion.

## Fully Custom Priors

Fully custom priors are useful when you have a specific substantive reason for the scale of the effect, heterogeneity, or moderator coefficients.
Custom priors are created with `prior()` and related helper functions.

```{r custom-effect-priors, eval = has_bcg}
bcg_custom_priors <- brma(
  yi                  = yi,
  vi                  = vi,
  data                = bcg,
  measure             = "RR",
  prior_effect        = prior(
    "normal",
    parameters = list(mean = 0, sd = 0.5)
  ),
  prior_heterogeneity = prior(
    "normal",
    parameters = list(mean = 0, sd = 0.25),
    truncation = list(lower = 0, upper = Inf)
  ),
  only_priors         = TRUE
)

print_prior(bcg_custom_priors, parameter = "mu")
print_prior(bcg_custom_priors, parameter = "tau")
```

For meta-regression, `prior_mods` controls the priors on moderator coefficients.
A single prior can be used for all continuous moderators.
A named list can be used when different predictors need different priors.

```{r custom-regression-priors}
Kroupova2021_continuous_priors <- brma(
  yi          = yi,
  vi          = vi,
  data        = Kroupova2021,
  measure     = "ZCOR",
  mods        = ~ sample_size,
  prior_mods  = prior(
    "normal",
    parameters = list(mean = 0, sd = 0.10)
  ),
  only_priors = TRUE
)

print_prior(Kroupova2021_continuous_priors, parameter_mods = "sample_size")
```

## Priors in Bayesian Model Averaging

The examples above use `brma()` because it is easier to understand priors in one model first.
`BMA()` and `RoBMA()` then use these priors inside Bayesian model averaging.
They average across models that include or exclude the effect, heterogeneity, and moderators.
This is represented with spike-and-slab priors.
The spike is a point prior at zero.
The slab is the continuous prior used when the parameter is included.

```{r bma-priors, eval = has_bcg}
bcg_bma_priors <- BMA(
  yi          = yi,
  vi          = vi,
  data        = bcg,
  measure     = "RR",
  only_priors = TRUE
)

print_prior(bcg_bma_priors, parameter = "mu")
print_prior(bcg_bma_priors, parameter = "tau")
```

In `BMA()` and `RoBMA()`, the package includes the null hypotheses unless you remove them.
For example, `prior_effect_null = NULL` means "do not include the no-effect model".
This is useful when you want estimation under an assumed effect rather than a test of effect versus no effect.

```{r bma-no-null, eval = has_bcg}
bcg_estimation_priors <- BMA(
  yi                = yi,
  vi                = vi,
  data              = bcg,
  measure           = "RR",
  prior_effect_null = NULL,
  only_priors       = TRUE
)

print_prior(bcg_estimation_priors, parameter = "mu")
```

## Publication-Bias Priors

Publication-bias priors describe which bias-adjustment models are included in the ensemble.
They are separate from the effect and heterogeneity priors.

The default `RoBMA()` setting is `model_type = "PSMA"`.
It averages across a no-bias model, several weight-function models, PET, and PEESE [@maier2020robust; @bartos2021no].
PET and PEESE adjust for small-study effects [@stanley2014meta].
Weight-function models adjust for selection across p-value regions [@vevea1995general].

```{r publication-bias-default, eval = has_bcg}
bcg_bias_priors <- RoBMA(
  yi          = yi,
  vi          = vi,
  data        = bcg,
  measure     = "RR",
  only_priors = TRUE
)

print_prior(bcg_bias_priors, parameter = "bias")
```

The preset can be changed with `model_type`.

```{r publication-bias-presets, eval = has_bcg}
bcg_pp_priors <- RoBMA(
  yi          = yi,
  vi          = vi,
  data        = bcg,
  measure     = "RR",
  model_type  = "PP",
  only_priors = TRUE
)

print_prior(bcg_pp_priors, parameter = "bias")
```

The most common choices are:

| `model_type` | Meaning |
|:--|:--|
| `"PSMA"` | default RoBMA-PSMA ensemble: weight functions plus PET-PEESE |
| `"PP"` | PET-PEESE only |
| `"6w"` | six weight-function models |
| `"2w"` | two weight-function models |

You can also supply a custom publication-bias prior.
The following example keeps the no-bias model and adds one one-sided weight function and one PET model.

```{r publication-bias-custom, eval = has_bcg}
bcg_custom_bias_priors <- RoBMA(
  yi              = yi,
  vi              = vi,
  data            = bcg,
  measure         = "RR",
  prior_bias_null = prior_none(),
  prior_bias      = list(
    prior_weightfunction(
      "one.sided",
      parameters = list(
        steps = c(0.025, 0.05),
        alpha = c(1, 1, 1)
      )
    ),
    prior_PET(
      "cauchy",
      parameters = list(location = 0, scale = 1),
      truncation = list(lower = 0, upper = Inf)
    )
  ),
  only_priors     = TRUE
)

print_prior(bcg_custom_bias_priors, parameter = "bias")
```

Custom bias priors are powerful, but they also change the model space being averaged over.
They should be chosen before looking at the results, and they should match the publication process that is plausible in the field.

## Practical Advice

Start with the default priors and always set `measure` carefully.
Transform correlations to Fisher's z and use `measure = "ZCOR"`.
Use `measure = "GEN"` with `ni` or `prior_unit_information_sd` when the effect size is not on a standardized scale.
Use informed priors when a relevant empirical prior exists for the research area.
Use custom priors when you can explain why the chosen scale is plausible.

For reporting, state the effect-size measure, whether the priors were default, informed, or custom, whether `rescale_priors` or `prior_unit_information_sd` was used, and which publication-bias ensemble was fitted.
This makes the Bayesian analysis easier to understand and easier to reproduce.
