---
title: "Multivariate and Multilevel Meta-Analysis"
author: "František Bartoš"
date: "22nd of June 2026"
output:
  rmarkdown::html_vignette:
    self_contained: yes
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Multivariate and Multilevel Meta-Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---

```{r child = "_vignette-nowrap.md", echo = FALSE, eval = TRUE}
```

```{r setup, include = FALSE}
source("vignette-cache.R", local = knitr::knit_global())

cached_fits <- vignette_cache(
  name     = "v14-metafor-parity-multivariate",
  objects  = c("fit1_brma", "fit2_brma", "fit3_brma", "fit4_brma"),
  packages = c("metafor", "metadat", "BayesTools")
)

knitr::opts_chunk$set(
  collapse    = TRUE,
  comment     = "#>",
  eval        = vignette_cache_eval(cached_fits),
  message     = FALSE,
  warning     = FALSE,
  fig.width   = 7,
  fig.height  = 3.5,
  dev         = "png",
  fig.retina  = 3
)
if (.Platform$OS.type == "windows") {
  knitr::opts_chunk$set(dev.args = list(type = "cairo"))
}
```

```{r load-models, include = FALSE}
library("metafor")
library("RoBMA")

vignette_cache_load(cached_fits)
```

```{r precompute-models, include = FALSE, eval = FALSE}
library("metafor")
library("RoBMA")

data("dat.konstantopoulos2011", package = "metadat")
dat_konstantopoulos <- dat.konstantopoulos2011

fit1_brma <- brma.mv(
  yi = yi, V = vi, measure = "SMD",
  random = ~ cs(school | district),
  data = dat_konstantopoulos, seed = 1,
  chains = 4, adapt = 2000, burnin = 5000, sample = 30000,
  parallel = TRUE, autofit = TRUE
)

data("dat.assink2016", package = "metadat")
dat_assink <- dat.assink2016
V_assink   <- metafor::vcalc(
  vi, cluster = study, type = deltype, obs = esid,
  rho = c(0.7, 0.5), data = dat_assink
)
V_assink[lower.tri(V_assink)] <- t(V_assink)[lower.tri(V_assink)]

fit2_brma <- brma.mv(
  yi = yi, V = V_assink, measure = "SMD",
  random = ~ 1 | study / esid,
  data = dat_assink, seed = 2,
  adapt = 5000, burnin = 10000, sample = 20000
)

data("dat.ishak2007", package = "metadat")
dat_ishak <- dat.ishak2007
dat_ishak <- reshape(
  dat_ishak,
  direction = "long",
  idvar     = "study",
  v.names   = c("yi", "vi"),
  varying   = list(c(2, 4, 6, 8), c(3, 5, 7, 9))
)
dat_ishak <- dat_ishak[order(dat_ishak$study, dat_ishak$time), ]
dat_ishak <- dat_ishak[!is.na(dat_ishak$yi), ]
rownames(dat_ishak) <- NULL
dat_ishak$time_factor <- factor(dat_ishak$time)
V_ishak <- metafor::vcalc(
  vi, cluster = study, time1 = time,
  phi = 0.97, data = dat_ishak
)
V_ishak[lower.tri(V_ishak)] <- t(V_ishak)[lower.tri(V_ishak)]
ishak_unit_information_sd <- estimate_unit_information_sd(sei = sqrt(14.3), ni = 15)
ishak_time_prior <- BayesTools::prior_factor(
  distribution = "normal",
  parameters   = list(mean = 0, sd = ishak_unit_information_sd),
  contrast     = "independent"
)

fit3_brma <- brma.mv(
  yi = yi, V = V_ishak, measure = "GEN",
  mods = ~ 0 + time_factor,
  random = ~ har(time | study),
  prior_effect              = NULL,
  prior_mods                = ishak_time_prior,
  prior_unit_information_sd = ishak_unit_information_sd,
  data = dat_ishak, seed = 3,
  chains = 4, adapt = 5000, burnin = 10000, sample = 30000,
  parallel = TRUE, autofit = TRUE
)

data("dat.begg1989", package = "metadat")
dat_begg <- dat.begg1989
dat_begg$trt <- relevel(factor(dat_begg$trt), ref = "CMO")
dat_begg$ni  <- as.numeric(round(with(dat_begg, yi * (1 - yi) / sei^2)))
begg_unit_information_sd <- estimate_unit_information_sd(sei = dat_begg$sei, ni = dat_begg$ni)
sd_prior <- BayesTools::prior(
  distribution = "normal",
  parameters = list(mean = 0, sd = begg_unit_information_sd),
  truncation = list(lower = 0, upper = Inf)
)
rho_zero <- BayesTools::prior(
  distribution = "spike",
  parameters = list(location = 0)
)
cs_zero <- BayesTools::random_covariance(
  rho       = rho_zero,
  rho_scale = "rho"
)

fit4_brma <- brma.mv(
  yi = yi, V = vi, ni = ni, measure = "GEN",
  mods = ~ trt,
  random = list(
    study     = ~ 1 | study,
    treatment = ~ cs(trt | study)
  ),
  prior_heterogeneity = BayesTools::prior_random(
    study     = BayesTools::random_block(sd = sd_prior),
    treatment = BayesTools::random_block(
      sd         = sd_prior,
      covariance = cs_zero
    )
  ),
  prior_unit_information_sd = begg_unit_information_sd,
  data = dat_begg, seed = 4
)

vignette_cache_save(cached_fits)
```

This vignette is a starting point for fitting multivariate and multilevel meta-analytic models with the `RoBMA` R package [@RoBMA].
We walk through four datasets from the `metadat` package [@metadat], showing the `brma.mv()` call alongside its `metafor::rma.mv()` counterpart [@metafor].
The examples move from a simple three-level model to known sampling covariance matrices, temporal correlation structures, and two simultaneous random structures.

The first two examples rely on default prior distributions and the last two examples rely on unit-information-scaled prior distributions; see the [*Prior Distributions*](v01-prior-distributions.html) vignette for more details.
For generic summaries, plots, and diagnostics that also apply to `brma.mv()` objects, see the [*Bayesian Meta-Analysis*](v02-bayesian-meta-analysis.html) vignette.

## Three-Level Model

The `dat.konstantopoulos2011` dataset from the `metadat` package contains 56 standardized mean differences from studies of modified school calendars and student achievement.
Positive values indicate higher achievement for students on modified calendars, and schools are clustered within districts.
Effect sizes `yi` and sampling variances `vi` are already provided.

```{r data-konstantopoulos}
data("dat.konstantopoulos2011", package = "metadat")
dat_konstantopoulos <- dat.konstantopoulos2011
head(dat_konstantopoulos)
```

The `metafor` package fit uses a compound-symmetric random effect for schools within districts.
We set `struct = "CS"` explicitly so that the fitted random structure is visible in the call.
The shorter `brma(..., cluster = district)` syntax for this special case is shown in the [*Multilevel Meta-Analysis*](v10-metafor-parity-multilevel.html) vignette.

```{r fit1-metafor}
fit1_metafor <- metafor::rma.mv(
  yi, vi,
  random = ~ school | district,
  struct = "CS",
  data = dat_konstantopoulos
)
fit1_metafor
```

The matching `brma.mv()` call passes the sampling variances through `V` and uses the `cs()` random-effect formula tag.
The `cs(school | district)` term estimates one shared heterogeneity standard deviation and one within-district correlation for the school effects.

```{r fit1-brma, eval = FALSE}
fit1_brma <- brma.mv(
  yi = yi, V = vi, measure = "SMD",
  random = ~ cs(school | district),
  data = dat_konstantopoulos, seed = 1,
  chains = 4, adapt = 2000, burnin = 5000, sample = 30000,
  parallel = TRUE, autofit = TRUE
)
```

`summary()` reports the pooled effect and the random-effect parameters monitored by the fitted Bayesian model.
Here `sd(shared | district)` corresponds to the `metafor` package estimate of total heterogeneity and `rho(district)` corresponds to the compound-symmetric correlation.

```{r fit1-brma-summary}
summary(fit1_brma, include_mcmc_diagnostics = FALSE)
```

## Known Sampling Covariance

The `dat.assink2016` dataset from the `metadat` package contains standardized mean differences from 17 studies on recidivism and mental health among delinquent juveniles.
Most studies contributed multiple estimates for different delinquent behaviors, so the example needs both a working sampling covariance matrix and random effect specification.
The `metafor::vcalc()` call constructs a working sampling covariance matrix from the effect-size variances, study labels, delinquency type, and effect-size identifiers. We copy its upper triangle to the lower triangle because floating-point evaluation can otherwise leave the two triangles microscopically different.

```{r data-assink}
data("dat.assink2016", package = "metadat")
dat_assink <- dat.assink2016
V_assink   <- metafor::vcalc(
  vi, cluster = study, type = deltype, obs = esid,
  rho = c(0.7, 0.5), data = dat_assink
)
V_assink[lower.tri(V_assink)] <- t(V_assink)[lower.tri(V_assink)]
head(dat_assink)
```

The `metafor` package model uses the full `V` matrix and a nested random intercept for studies and effect sizes within studies.

```{r fit2-metafor}
fit2_metafor <- metafor::rma.mv(
  yi, V_assink,
  random = ~ 1 | study / esid,
  data = dat_assink
)
fit2_metafor
```

The `brma.mv()` call mirrors the same inputs: `V` is the known sampling covariance matrix and `random` describes the additional heterogeneity.
The nested variance allocation can mix slowly, so this example uses a longer MCMC run.

```{r fit2-brma, eval = FALSE}
fit2_brma <- brma.mv(
  yi = yi, V = V_assink, measure = "SMD",
  random = ~ 1 | study / esid,
  data = dat_assink, seed = 2,
  adapt = 5000, burnin = 10000, sample = 20000
)
```

```{r fit2-brma-summary}
summary(fit2_brma, include_mcmc_diagnostics = FALSE)
```

The known `V` matrix and the random-effects formula address different sources of dependence.
The `V` matrix encodes sampling-error dependence that is treated as known, while the random-effects formula estimates residual heterogeneity across the levels named in `random`.

```{r fit2-brma-heterogeneity}
summary_heterogeneity(fit2_brma)
```

The first heterogeneity table summarizes the shared `study/esid` heterogeneity allocation, including the total heterogeneity and variance fractions; the following tables summarize the block-specific standard deviations.

## Heterogeneous AR(1) Structure

The `dat.ishak2007` dataset from the `metadat` package contains results from 46 studies of deep-brain stimulation for Parkinson's disease.
Negative mean differences indicate improvements in UPDRS motor-function scores, and repeated follow-up measurements within studies motivate a temporal correlation model.
The data are first reshaped to long format, and `metafor::vcalc()` uses a working autocorrelation of `0.97` for the sampling errors. We again copy the upper triangle to construct the exactly symmetric covariance matrix required by `brma.mv()`.

```{r data-ishak}
data("dat.ishak2007", package = "metadat")
dat_ishak <- dat.ishak2007
dat_ishak <- reshape(
  dat_ishak,
  direction = "long",
  idvar     = "study",
  v.names   = c("yi", "vi"),
  varying   = list(c(2, 4, 6, 8), c(3, 5, 7, 9))
)
dat_ishak <- dat_ishak[order(dat_ishak$study, dat_ishak$time), ]
dat_ishak <- dat_ishak[!is.na(dat_ishak$yi), ]
rownames(dat_ishak) <- NULL
dat_ishak$time_factor <- factor(dat_ishak$time)
V_ishak <- metafor::vcalc(
  vi, cluster = study, time1 = time,
  phi = 0.97, data = dat_ishak
)
V_ishak[lower.tri(V_ishak)] <- t(V_ishak)[lower.tri(V_ishak)]
head(dat_ishak)
```

The `metafor` package uses `struct = "HAR"` for a heterogeneous AR(1) true-effect structure over time within studies.
The fixed effects estimate one mean difference at each time point, with the intercept suppressed.

```{r fit3-metafor}
fit3_metafor <- metafor::rma.mv(
  yi, V_ishak,
  mods = ~ 0 + time_factor,
  random = ~ time | study,
  struct = "HAR",
  data = dat_ishak
)
fit3_metafor
```

The `brma.mv()` call uses the BayesTools `har()` wrapper in the random-effects formula.
Because this outcome is a generic continuous outcome and sample sizes are not reported in the source table, we use one of the primary-study sample sizes to estimate the unit-information scale.
The Alegret (2001) row reports a 3-month sampling variance of `14.3`, and the primary study reports `15` patients.
We compute the corresponding unit-information standard deviation with `estimate_unit_information_sd()` and use it directly in the prior distribution specification.
The independent factor prior distribution gives each time point its own location coefficient on this UPDRS scale.

```{r fit3-brma, eval = FALSE}
ishak_unit_information_sd <- estimate_unit_information_sd(sei = sqrt(14.3), ni = 15)

fit3_brma <- brma.mv(
  yi = yi, V = V_ishak, measure = "GEN",
  mods = ~ 0 + time_factor,
  random = ~ har(time | study),
  prior_effect              = NULL,
  prior_mods                = BayesTools::prior_factor(
    distribution = "normal",
    parameters   = list(mean = 0, sd = ishak_unit_information_sd),
    contrast     = "independent"
  ),
  prior_unit_information_sd = ishak_unit_information_sd,
  data = dat_ishak, seed = 3,
  chains = 4, adapt = 5000, burnin = 10000, sample = 30000,
  parallel = TRUE, autofit = TRUE
)
```

```{r fit3-brma-summary}
summary(fit3_brma, include_mcmc_diagnostics = FALSE)
```

The `HAR` structure lets the heterogeneity differ by time point while estimating one autocorrelation across neighboring time points.
This separates the temporal structure of the true effects from the working temporal dependence already encoded in `V_ishak`.

## Study and Treatment Random Effects

The `dat.begg1989` dataset from the `metadat` package contains 20 estimates of two-year disease-free survival rates for acute nonlymphocytic leukemia.
The studies compare allogeneic bone-marrow transplantation (BMT) and conventional chemotherapy (CMO), with both controlled two-arm studies and uncontrolled one-arm studies.
The example uses the reported survival rates and their sampling variances directly on the raw rate scale.

```{r data-begg}
data("dat.begg1989", package = "metadat")
dat_begg <- dat.begg1989
dat_begg$trt <- relevel(factor(dat_begg$trt), ref = "CMO")
dat_begg$ni  <- as.numeric(round(with(dat_begg, yi * (1 - yi) / sei^2)))
head(dat_begg)
```

The `metafor` package model combines a study intercept with a treatment-specific random effect.
The study intercept captures heterogeneity in baseline survival.
The treatment-specific component allows the BMT and CMO estimates to vary within studies, with the treatment-level correlation fixed at zero by `rho = 0`.

```{r fit4-metafor}
fit4_metafor <- metafor::rma.mv(
  yi, vi,
  mods = ~ trt,
  random = list(
    ~ 1 | study,
    ~ trt | study
  ),
  struct = "CS",
  rho    = 0,
  data   = dat_begg,
  method = "ML"
)
fit4_metafor
```

The `brma.mv()` sets names for the two random structures explicitly.
Because the source table reports survival rates and standard errors but not sample sizes, we recover effective sample sizes and use them to scale the prior distributions.
The `prior_random()` call specifies the study and treatment blocks directly because it fixes the treatment-level correlation at zero.

```{r fit4-brma, eval = FALSE}
begg_unit_information_sd <- estimate_unit_information_sd(sei = dat_begg$sei, ni = dat_begg$ni)

sd_prior <- BayesTools::prior(
  distribution = "normal",
  parameters = list(mean = 0, sd = begg_unit_information_sd),
  truncation = list(lower = 0, upper = Inf)
)
rho_zero <- BayesTools::prior(
  distribution = "spike",
  parameters = list(location = 0)
)
cs_zero <- BayesTools::random_covariance(
  rho       = rho_zero,
  rho_scale = "rho"
)

fit4_brma <- brma.mv(
  yi = yi, V = vi, ni = ni, measure = "GEN",
  mods = ~ trt,
  random = list(
    study     = ~ 1 | study,
    treatment = ~ cs(trt | study)
  ),
  prior_heterogeneity = BayesTools::prior_random(
    study     = BayesTools::random_block(sd = sd_prior),
    treatment = BayesTools::random_block(
      sd         = sd_prior,
      covariance = cs_zero
    )
  ),
  prior_unit_information_sd = begg_unit_information_sd,
  data = dat_begg, seed = 4
)
```

```{r fit4-brma-summary}
summary(fit4_brma, include_mcmc_diagnostics = FALSE)
```

The study component captures baseline heterogeneity across studies, while the treatment component captures additional treatment-specific variation within studies.

```{r fit4-brma-heterogeneity}
summary_heterogeneity(fit4_brma)
```


## General Considerations and Reporting

**Report both covariance layers.** State how sampling covariance `V` was constructed and separately describe every estimated or known-`R` random component. Also report the effective known-`V` backend shown by `summary()`.

**Name the downstream target.** Distinguish conditional fitted effects, marginal new-effect predictions, Schur-conditional estimate diagnostics, marginal GLS diagnostics, and the full joint marginal likelihood. For LOO, report Pareto-k diagnostics and note that dependency-block deletion is unavailable.

**Make future designs explicit.** Report the covariate and random-component levels used for prediction, and provide `V_new` when predicting future observed estimates. Unseen known-`R` levels require a future covariance interface and are therefore rejected.


## References
