---
title: "Multilevel Robust Bayesian Model-Averaged Meta-Regression"
author: "František Bartoš"
date: "2025"
output:
  rmarkdown::html_vignette:
    self_contained: yes
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa.csl
vignette: >
  %\VignetteIndexEntry{Multilevel Robust Bayesian Model-Averaged Meta-Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
is_check <- ("CheckExEnv" %in% search()) ||
              any(c("_R_CHECK_TIMINGS_", "_R_CHECK_LICENSE_") %in% names(Sys.getenv())) ||
              !file.exists("../models/MultilevelRoBMARegression/zfit_Havrankova2025.RDS")
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = !is_check,
  dev       = "png")
if(.Platform$OS.type == "windows"){
  knitr::opts_chunk$set(dev.args = list(type = "cairo"))
}
```

```{r include = FALSE}
library(RoBMA)
zfit_reg <- readRDS(file = "../models/MultilevelRoBMARegression/zfit_Havrankova2025.RDS")
fit_reg  <- zfit_reg
class(fit_reg) <- class(fit_reg)[!class(fit_reg) %in% "zcurve_RoBMA"]
```

```{r include = FALSE, eval = FALSE}
# R package version updating
library(RoBMA)
data("Havrankova2025", package = "RoBMA")

# Prior scaling
fit_fe        <- metafor::rma(yi = y, sei = se, data = Havrankova2025, method = "FE")
unti_scale    <- fit_fe$se * sqrt(sum(Havrankova2025$N))
prior_scale   <- unti_scale * 0.5

df_reg  <- data.frame(
  y               = Havrankova2025$y,
  se              = Havrankova2025$se, 
  facing_customer = Havrankova2025$facing_customer, 
  cluster_id      = Havrankova2025$study_id
)

fit_reg <- RoBMA.reg(
  ~ facing_customer, 
  cluster        = df_reg$cluster_id,
  data           = df_reg, 
  rescale_priors = prior_scale,
  prior_scale = "none", transformation = "none",
  algorithm = "ss", sample = 20000, burnin = 10000, adapt = 10000, 
  thin = 5, parallel = TRUE, autofit = FALSE, seed = 1)
)

zfit_reg    <- as_zcurve(fit_reg)
saveRDS(zfit_reg, file = "../models/MultilevelRoBMARegression/zfit_Havrankova2025.RDS", compress = "xz")
```

**This vignette accompanies the manuscript [Robust Bayesian Multilevel Meta-Analysis: Adjusting for Publication Bias in the Presence of Dependent Effect Sizes](https://doi.org/10.31234/osf.io/9tgp2_v1) preprinted at *PsyArXiv* [@bartos2025robust].**

This vignette reproduces the second example from the manuscript, re-analyzing data from @havrankova2025beauty on the relationship between beauty and professional success (1,159 effect sizes from 67 studies). 
We investigate whether the effect depends on the type of customer contact ("no", "some", or "direct"). 
For the first example describing a publication bias-adjusted multilevel meta-analysis see [Multilevel Robust Bayesian Meta-Analysis](MultilevelRoBMA.html).

### Rescaling Priors for Non-Standardized Effect Sizes

The effect sizes in this dataset are not standardized (percent increase in earnings with one-standard deviation increase in beauty). 
Therefore, we must rescale the default prior distributions to avoid specifying overly narrow or wide priors. 
Overly narrow priors would make it difficult to find evidence for the effect (as both models would predict effects very close to zero--their predictions would overlap substantially), 
while overly wide priors would bias the results towards the null hypothesis (as alternative models would predict unrealistically large effects--their predictions would never be supported by the data).

We rescale the priors by matching the default prior scale to the unit information scale of the data [@rover2021weakly; @mulder2024bayesian]. 
The unit information scale can be approximated by the standard error of a fixed-effect model multiplied by the square root of the total sample size. 
Since the default prior for standardized mean differences has a standard deviation of one (corresponding to half the unit information), we multiply the unit information scale by 0.5. 

```{r}
library(RoBMA)
data("Havrankova2025", package = "RoBMA")

fit_fe        <- metafor::rma(yi = y, sei = se, data = Havrankova2025, method = "FE")
unti_scale    <- fit_fe$se * sqrt(sum(Havrankova2025$N))
prior_scale   <- unti_scale * 0.5
```

The resulting `prior_scale`, in our example approximately 26, can be used in the `rescale_priors` argument in the fitting functions of the `RoBMA` package,

```{r, eval = FALSE}
df_reg  <- data.frame(
  y               = Havrankova2025$y,
  se              = Havrankova2025$se, 
  facing_customer = Havrankova2025$facing_customer, 
  cluster_id      = Havrankova2025$study_id
)

fit_reg <- RoBMA.reg(
  ~ facing_customer, 
  cluster        = df_reg$cluster_id,
  data           = df_reg, 
  rescale_priors = prior_scale,
  prior_scale    = "none", transformation = "none",
  algorithm = "ss", sample = 20000, burnin = 10000, adapt = 10000, 
  thin = 5, parallel = TRUE, autofit = FALSE, seed = 1)
)
```

where we first specify the input `data.frame` (with `y` and `se` denoting the effect size and standard error inputs) and `prior_scale = "none"` and `transformation = "none"` arguments disable prior distributions for standardized effect sizes. 
We also notably increase the number of adaptation, burn-in, and sampling iterations to ensure convergence of the more complex model (and use the `algorithm = "ss"` options since the bridge-sampling algorithm is unfeasible for multilevel models).

*Note that the meta-regression model can take more than an hour to run with parallel processing enabled.*

### Interpreting the Results

The `summary` and `marginal_summary` functions provide the overall model estimates and the estimated marginal means.

```{r}
summary(fit_reg)
```

Using the rescaled prior distributions, the multilevel RoBMA meta-regression finds extreme evidence for the presence of the average effect, moderation by the degree of consumer contact, between-study heterogeneity, and publication bias (all BFs > 10^6). 
The model-averaged average effect estimate of mu = 3.0, 95% CI [1.9, 4.2], is accompanied by considerable overall heterogeneity tau = 4.5, 95% CI [3.9, 5.3], 
and cluster-level allocation close to 1, rho = 0.82, 95% CI [0.75, 0.88], indicating a high degree of similarity of estimates from the same study.

The summary function also warns us about a large R-hat. Inspection of the parameter-specific MCMC diagnostics via `summary(fit_reg, type = "diagnostics")` shows that while some R-hats are inflated, most parameters were sampled to an acceptable level. The main source of issue seems to be the moderator coefficients `facing_customer`. We visually assess the chains `diagnostics_trace(fit_reg, "facing_customer")` but do not notice substantial issues.

The average effect is moderated by the degree of consumer contact, to examine the effect at each level of the moderator we can use the `marginal_summary()` function.

```{r}
marginal_summary(fit_reg)
```

The estimated marginal means at each level of the moderator reveals moderate evidence for the absence of the effect for no consumer contact jobs mu = 0.9, 95% CI [-0.5, 2.5], BF10 = 0.11, 
and extreme evidence for jobs with some consumer contact mu = 4.1, 95% CI [2.6, 5.5], BF10 > 10^6 and jobs with direct consumer contact mu = 4.1, 95% CI [2.7, 5.5], BF10 > 10^6.

### Visual Fit Assessment

We can visualize evalate the model fit using the meta-analytic z-curve plot [@bartos2025zcurve] (also see [Z-Curve Publication Bias Diagnostics](ZCurveDiagnostics.html) vignette for more details).
The z-curve plot shows the distribution of the observed $z$-statistics (computed as the effect size / standard error), with dotted red horizontal lines highlighting the typical steps on which publication bias operates 
($z = \pm 1.65$ and $z = \pm 1.96$ corresponding to statistically significant and marginally significant $p$-values, and $z = 0$ corresponding to the selection of the direction of the effect).

```{r, eval = FALSE}
zfit_reg <- as_zcurve(fit_reg)
```
```{r, fig.width = 6, fig.height = 4}
par(mar = c(4,4,0,0))
plot(zfit_reg, by.hist = 0.25, plot_extrapolation = FALSE, from = -4, to = 8)
```

We can notice clear discontinuities corresponding to the selection on the direction of the effect and marginal significance. 
The posterior predictive density from the multilevel RoBMA meta-regression indicates that the RoBMA meta-regression model approximates the observed data reasonably well, 
capturing the discontinuities and the proportion of statistically significant results. This reassures us that the model provides a good fit to the data while adjusting for publication bias.

### References
